---
title: "Distribution families, network types, and data preparation"
author: "Shahryar Minhas & Peter Hoff"
description: "Use Normal and Binomial outcomes, symmetric and bipartite networks, dynamic influence covariates, and array builders."
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Distribution families, network types, and data preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = TRUE
)
library(sir)
```

The SIR framework handles a range of outcome types, network structures, and covariate configurations beyond the Poisson directed-network case in the overview. This vignette covers the distribution families, symmetric and bipartite networks, dynamic (time-varying) influence covariates, and the data-preparation utilities. For basic fitting see `vignette("sir_overview")`.

## Distribution families

Every example below recovers a known influence operator so you can see the estimator work for each family and network type. The `sim_sir()` examples generate the data directly; the bipartite and dynamic-`W` sections instead build small networks by hand from a known operator --- `sim_sir()` makes only square, one-mode data, so those shapes have to be constructed manually --- and then check that the fit recovers it.

### Normal (continuous outcomes)

For continuous relational data (sentiment scores, net flows, log trade ratios) the Normal family uses an identity link. The influence parameters are interpreted exactly as in the Poisson case: $\alpha$ and $\beta$ identify which covariates account for sender and receiver influence, with effects entering the linear predictor.

```{r normal}
dn  <- sim_sir(m = 14, T_len = 90, p = 2, q = 2, family = "normal",
               gain = 0.9, seed = 11)
fit_norm <- sir(dn$Y, W = dn$W, X = dn$X, Z = dn$Z,
                family = "normal", calc_se = FALSE, seed = 1)
target_norm <- c(dn$theta, dn$alpha[-1], dn$beta)
data.frame(
	term = names(coef(fit_norm)),
	estimate = round(unname(coef(fit_norm)), 3),
	target = round(target_norm, 3),
	abs_error = round(abs(unname(coef(fit_norm)) - target_norm), 3),
	row.names = NULL
)
```

The sender block shows only `(alphaW) W2` because the first sender weight is fixed at 1 for identification (the $\alpha_1 = 1$ anchor) and so is not estimated; the receiver block shows both `(betaW) W1` and `(betaW) W2`. All five estimated coefficients land close to their generating values with the correct signs, so the Normal family recovers sender and receiver influence just as the Poisson fit in the overview did.

On the response scale, `predict()` returns fitted means directly. Because we simulated a strong influence signal (`gain = 0.9`), those fitted means track the observed network closely --- the correlation between fitted and observed is high, so the model is capturing the real structure, not just the average level:

```{r normal-response}
mu_norm <- predict(fit_norm)                 # fitted means (identity link)
round(cor(c(mu_norm), c(dn$Y), use = "complete.obs"), 3)
```

### Binomial (binary outcomes)

For binary relational outcomes (presence/absence of a tie, conflict onset) the Binomial family uses a logit link. The full bilinear contribution $(A X_t B^\top)_{ij}$ enters the log-odds, so an individual $\alpha$ or $\beta$ coefficient is not a standalone odds-ratio term; its effect depends on `X`, the other influence side, and the covariate scale.

```{r binomial}
db  <- sim_sir(m = 16, T_len = 130, p = 2, q = 2, family = "binomial",
               alpha = c(1, 0.8), beta = c(0.7, -0.6), seed = 12)
fit_bin <- sir(db$Y, W = db$W, X = db$X, Z = db$Z,
               family = "binomial", calc_se = FALSE, seed = 1)
target_bin <- c(db$theta, db$alpha[-1], db$beta)
data.frame(
	term = names(coef(fit_bin)),
	estimate = round(unname(coef(fit_bin)), 3),
	target = round(target_bin, 3),
	abs_error = round(abs(unname(coef(fit_bin)) - target_bin), 3),
	row.names = NULL
)
```

Despite the binary outcome, every coefficient lands within about 0.02 of its generating value (see `abs_error`) --- the influence weights recover as cleanly as the direct effects.

`predict()` returns fitted tie probabilities by default. Pass `type = "link"` instead if you want the log-odds (linear-predictor) scale; apply `plogis()` to those values to convert back to probabilities. A more telling check than the overall tie rate is whether the model *discriminates* --- does it assign higher probabilities to dyads that actually tied than to those that did not?

```{r binomial-prob}
p_hat <- predict(fit_bin)                    # fitted tie probabilities
c(mean_prob_at_observed_ties     = round(mean(p_hat[db$Y == 1], na.rm = TRUE), 3),
  mean_prob_at_observed_non_ties = round(mean(p_hat[db$Y == 0], na.rm = TRUE), 3))
```

The model assigns a clearly higher probability to dyads that tied than to those that did not, so the influence structure carries real predictive signal --- not just the right marginal rate.

## Symmetric (undirected) networks

For undirected networks where $y_{i,j} = y_{j,i}$ (alliances, trade, co-sponsorship) set `symmetric = TRUE`. The sender and receiver roles collapse into one operator, $A = B = \sum_k \gamma_k W_k$, so the bilinear term becomes the **quadratic form** $A X A'$ and there is a single set of influence weights $\gamma$, estimated freely with no fixed anchor. `sim_sir(symmetric = TRUE)` generates a matching undirected process, so we can check recovery of a known $\gamma$:

```{r symmetric}
ds <- sim_sir(m = 14, T_len = 70, p = 3, q = 1, family = "normal",
              symmetric = TRUE, alpha = c(1.5, 0.9, -0.6), seed = 7)

fit_sym <- sir(ds$Y, W = ds$W, X = ds$X, Z = ds$Z,
               family = "normal", symmetric = TRUE, seed = 1)

data.frame(
	term       = names(coef(fit_sym)),
	estimate   = round(unname(coef(fit_sym)), 3),
	target     = round(c(ds$theta, ds$alpha), 3),
	se_cluster = round(sqrt(diag(vcov(fit_sym))), 3),
	row.names = NULL
)
cat("operator A == B:", isTRUE(all.equal(fit_sym$A, fit_sym$B)),
    "| diag(A) == 0:", max(abs(diag(fit_sym$A))) < 1e-10, "\n")
```

The estimates track the true shared weights (the `target` column, stored sim-side in `ds$alpha`), and the fitted network is exactly symmetric with a zero diagonal. Unlike the directed case there is no fixed anchor --- every $\gamma_k$ is estimated, including the first (target 1.5, not pinned at 1).

A few specifics for applied use: $A$ is identified only up to overall sign (since $A X A' = (-A) X (-A)'$), fixed so the largest $|\gamma_k|$ is positive --- match signs before pooling $\gamma$ across separate fits (`boot_sir()` already realigns its resamples). Inputs must be a square network with symmetric, zero-diagonal `W`; Poisson/Binomial `Y` must already be symmetric, and continuous `Y` is averaged across triangles. As for directed fits, `confint()` and `vcov()` default to the cluster-robust SE (see `vignette("sir_inference")`).

## Bipartite networks

When senders and receivers are **distinct populations** (a rectangular $Y$, e.g. countries $\to$ NGOs, or legislators co-sponsoring others' bills), bipartite structure is detected automatically and `fix_receiver = TRUE` is enforced.

**Influence flows on the sender side, so `W` describes the senders.** For an $n_1 \times n_2$ outcome, `W` must be $n_1 \times n_1 \times p$ (sender-by-sender), *not* keyed to the receivers. `X` matches `Y` at $n_1 \times n_2$. Exogenous covariates `Z` are optional and, like `X`, are keyed to the outcome --- supply them as $n_1 \times n_2 \times q \times T$ (the example below omits `Z` for brevity). Because `sim_sir()` generates only square networks, we build this rectangular example by hand from a known sender operator and confirm the fit recovers it.

```{r bipartite}
set.seed(7)
n1 <- 12; n2 <- 8; T_len <- 80; p <- 2
W_bp <- array(rnorm(n1 * n1 * p), dim = c(n1, n1, p))   # sender-by-sender
A_bp <- W_bp[, , 1] + 0.6 * W_bp[, , 2]                 # true sender operator (alpha = 1, 0.6)

Y_bp <- array(0, dim = c(n1, n2, T_len))
X_bp <- array(0, dim = c(n1, n2, T_len))
for (t in 2:T_len) {
	X_bp[, , t] <- Y_bp[, , t - 1] / (n1 - 1)            # lagged signal, /(n1-1) keeps the bilinear sum on scale
	Y_bp[, , t] <- A_bp %*% X_bp[, , t] + matrix(rnorm(n1 * n2, sd = 0.4), n1, n2)
}

fit_bp <- sir(Y_bp, W = W_bp, X = X_bp,
              family = "normal", fix_receiver = TRUE, calc_se = FALSE, seed = 1)
data.frame(
	bipartite        = fit_bp$bipartite,
	fix_receiver     = fit_bp$fix_receiver,
	senders          = fit_bp$n1,
	receivers        = fit_bp$n2,
	alphaW2_estimate = round(unname(fit_bp$alpha[2]), 3),
	alphaW2_target   = 0.6,
	row.names = NULL
)
```

The output confirms both that the call did the right thing structurally --- the rectangular outcome was recognized as bipartite and the receiver side was fixed --- and that the sender-side influence weight is recovered (the estimate lands near its true value of 0.6). The full-bilinear example below adds a separate receiver-side channel.

For a *square* array whose rows and columns are nonetheless distinct populations, set `bipartite = TRUE` explicitly so the receiver side is not estimated.

### Full bilinear bipartite (with `W_recv`)

The default above collapses `B` to the identity. To estimate the full bilinear model $A X B'$ for two-mode data --- with a *separate* receiver-side influence structure --- supply `W_recv`, an $n_2 \times n_2 \times p_2$ array. The sender influence `A` is then built from `W` and the receiver influence `B` from `W_recv`, fit by alternating GLM with `alpha_1 = 1` pinning the scale.

Here we simulate a two-mode network *with* genuine bilinear structure so the fit has something to recover. True sender weights are `alpha = (1, 0.6)` and receiver weights `beta = (0.8, -0.5)`:

```{r bipartite-full}
set.seed(204)
n1 <- 10; n2 <- 6; Tn <- 30
Wf  <- array(rnorm(n1 * n1 * 2), dim = c(n1, n1, 2))   # sender-by-sender
Wr  <- array(rnorm(n2 * n2 * 2), dim = c(n2, n2, 2))   # receiver-by-receiver
Xf  <- array(rnorm(n1 * n2 * Tn) / sqrt(n2), dim = c(n1, n2, Tn))
A_t <- Wf[, , 1] + 0.6 * Wf[, , 2]
B_t <- 0.8 * Wr[, , 1] - 0.5 * Wr[, , 2]
Yf  <- array(0, dim = c(n1, n2, Tn))
for (t in 1:Tn) Yf[, , t] <- A_t %*% Xf[, , t] %*% t(B_t) +
    matrix(rnorm(n1 * n2, sd = 0.3), n1, n2)

fit_full <- sir(Yf, W = Wf, X = Xf, W_recv = Wr,
                family = "normal", calc_se = FALSE, seed = 1)
data.frame(
	full_bilinear = isTRUE(fit_full$full_bilinear),
	converged = fit_full$convergence
)

data.frame(
	term = names(coef(fit_full)),
	estimate = round(unname(coef(fit_full)), 3),
	target = c(0.6, 0.8, -0.5),
	row.names = NULL
)
```

All three estimates land essentially on their generating values, so the full bilinear model recovers the separate sender and receiver influence channels at once --- the most demanding recovery in this vignette.

Analytic standard errors are not available on this path; use the delete-one-actor dyad jackknife for inference:

```{r bipartite-full-boot}
bj <- boot_sir(fit_full, type = "dyad", seed = 1)
round(confint(bj), 3)
```

All three jackknife intervals comfortably cover the true weights (0.6, 0.8, -0.5), giving the full-bilinear fit a clean inferential payoff.

## Dynamic influence covariates

Often the factors that mediate influence change over time --- alliances evolve, trade shifts. Supply `W` as a 4D array ($m \times m \times p \times T$). The influence *coefficients* are estimated as time-invariant, but the reconstructed operators $A_t$ vary with $t$ because the covariates do. Dynamic 4D `W` is supported with a fixed receiver side (`fix_receiver = TRUE`, so $B = I$ and only the sender weights $\boldsymbol{\alpha}$ are estimated). Below we let the covariates evolve gradually --- an AR(1) drift across periods, like alliances shifting slowly --- and check that the fit still recovers the shared $\alpha$:

```{r dynamic-w}
set.seed(11)
m <- 14; T_len <- 80; p <- 2

# influence covariates that drift gradually: AR(1) across periods (rho = 0.9)
W_dyn <- array(0, dim = c(m, m, p, T_len))
W_dyn[, , , 1] <- rnorm(m * m * p)
for (t in 2:T_len) {
	W_dyn[, , , t] <- 0.9 * W_dyn[, , , t - 1] +
		sqrt(1 - 0.9^2) * array(rnorm(m * m * p), dim = c(m, m, p))
}
A_dyn <- function(t) W_dyn[, , 1, t] + 0.7 * W_dyn[, , 2, t]   # true operator (alpha = 1, 0.7)

Y_dyn <- array(0, dim = c(m, m, T_len))
X_dyn <- array(0, dim = c(m, m, T_len))
for (t in seq_len(T_len)) {
	if (t > 1) {
		lag <- Y_dyn[, , t - 1]; lag[is.na(lag)] <- 0
		X_dyn[, , t] <- lag / (m - 1)                  # lagged signal
	}
	Y_dyn[, , t] <- A_dyn(t) %*% X_dyn[, , t] + matrix(rnorm(m * m, sd = 0.4), m, m)
	diag(Y_dyn[, , t]) <- NA
}

fit_dyn <- sir(Y_dyn, W = W_dyn, X = X_dyn, family = "normal",
               fix_receiver = TRUE, calc_se = FALSE, seed = 1)
data.frame(
	alphaW2_estimate = round(unname(fit_dyn$alpha[2]), 3),
	alphaW2_target   = 0.7,
	A_dimensions     = paste(dim(fit_dyn$A), collapse = " x "),
	dynamic_W        = fit_dyn$dynamic_W,
	row.names = NULL
)
```

The shared coefficient is recovered (the `alpha_2` estimate lands near its true 0.7), and `A` is returned as an $m \times m \times T$ array. Because the covariates drift gradually, so does the operator --- we can see this in the across-period correlation of the fitted $A_t$:

```{r dynamic-w-edges}
c(cor_A1_vs_A2 = round(cor(c(fit_dyn$A[, , 1]), c(fit_dyn$A[, , 2])), 2),
  cor_A1_vs_AT = round(cor(c(fit_dyn$A[, , 1]), c(fit_dyn$A[, , T_len])), 2))
```

Adjacent periods are highly correlated (the operator barely moves from one step to the next) while the first and last periods are nearly uncorrelated --- the influence structure drifts smoothly over time rather than staying fixed or jumping at random.

## Data preparation

### Converting edge lists to arrays

Network data often arrives as a long-format edge list (sender, receiver, time, value). `cast_array()` converts it to the arrays `sir()` expects.

```{r cast-array}
set.seed(4)
edge_list <- expand.grid(i = paste0("n", 1:5), j = paste0("n", 1:5), t = 1:3)
edge_list <- edge_list[edge_list$i != edge_list$j, ]
edge_list$conflict <- rpois(nrow(edge_list), lambda = 3)
edge_list <- edge_list[!(edge_list$i == "n1" & edge_list$j == "n2"), ]   # n1 -> n2 never observed

Y_from_el <- cast_array(edge_list, var = "conflict")
dim(Y_from_el)
Y_from_el[1:4, 1:4, 1]      # diagonal and the unobserved n1 -> n2 cell both default to 0
```

`cast_array()` lays the observed counts into the $m \times m \times T$ array `sir()` expects and fills any pair absent from the edge list with 0 --- here both the self-ties on the diagonal and the `n1 -> n2` dyad we dropped. `sir()` ignores the diagonal when fitting, so self-ties never enter the likelihood. Note that for count outcomes an absent pair becomes an observed 0, indistinguishable from a genuine zero count, so make sure your edge list includes the dyads you intend to treat as observed.

### Constructing relational covariates

`rel_covar()` builds main ($z_{ij}$), reciprocal ($z_{ji}$), and transitive ($\sum_k s_{ik} s_{kj}$, where $s_{ik}$ is the symmetrized tie between $i$ and $k$) effects from a base dyadic variable. These capture common relational mechanisms and drop straight into the exogenous covariate array $Z$.

```{r rel-covar}
set.seed(5)
trade   <- array(abs(rnorm(8 * 8 * 4)), dim = c(8, 8, 4))   # base dyadic variable
Z_trade <- rel_covar(trade, "trade")
dim(Z_trade)
dimnames(Z_trade)[[3]]
```

The output is $8 \times 8 \times 3 \times 4$: the third dimension holds the three layers just named (main, reciprocal, transitive) and the fourth carries the original four time slices.

### Simulating SIR data

`sim_sir()` (used throughout these vignettes) generates synthetic data from the SIR process --- useful for simulation studies, power analysis, and verifying parameter recovery under controlled conditions.

```{r sim-sir}
dat <- sim_sir(m = 10, T_len = 8, p = 2, q = 1, family = "poisson", seed = 42)

# component shapes returned by sim_sir()
sapply(dat[c("Y", "X", "W", "Z")], function(a) paste(dim(a), collapse = " x "))

# the generating coefficients a fit should recover
list(alpha = round(dat$alpha, 2), beta = round(dat$beta, 2), theta = round(dat$theta, 2))
```

`sim_sir()` returns the arrays in the shapes `sir()` expects, together with the generating coefficients `alpha`, `beta`, and `theta`. With `cast_array()` and `rel_covar()` to build `Y`/`X`/`Z` from your own edge lists and covariates, and `sim_sir()` to prototype and run power analyses, you have the full toolkit to take a SIR model from raw data to a fitted, interpretable result --- see `vignette("sir_overview")` and `vignette("sir_inference")` to put it to work.
