---
title: "Distribution families, network types, and data preparation"
author: "Shahryar Minhas & Peter Hoff"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Distribution families, network types, and data preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = FALSE
)
library(sir)
library(ggplot2)
```

The SIR framework accommodates a range of outcome types, network structures, and covariate configurations beyond the Poisson directed-network case presented in the overview. This vignette covers the available distribution families, symmetric and bipartite network structures, dynamic (time-varying) influence covariates, and data preparation utilities for converting common data formats into the arrays required by `sir()`. For basic model fitting see `vignette("sir_overview")`.

## Setup

We reuse the influence covariates and exogenous covariate from the overview.

```{r data}
set.seed(42)
m = 15; T_len = 8; p = 2

W = array(0, dim = c(m, m, p))
geo = matrix(runif(m * m), m, m); geo = (geo + t(geo)) / 2; diag(geo) = 0
W[,,1] = geo
groups = sample(1:3, m, replace = TRUE)
W[,,2] = outer(groups, groups, "==") * 1.0; diag(W[,,2]) = 0
dimnames(W) = list(paste0("n", 1:m), paste0("n", 1:m), c("proximity", "shared_group"))

Z = array(0, dim = c(m, m, 1, T_len))
dist_mat = matrix(rnorm(m * m), m, m); dist_mat = (dist_mat + t(dist_mat)) / 2
diag(dist_mat) = NA
for (t in 1:T_len) Z[,,1,t] = dist_mat
dimnames(Z)[[3]] = "distance"
```

## Distribution families

### Normal (continuous outcomes)

For continuous relational data such as trade volumes or sentiment scores, the Normal family uses an identity link. The interpretation of the influence parameters is the same as in the Poisson case: the $\alpha$ and $\beta$ coefficients identify which covariates account for sender and receiver influence patterns.

```{r normal}
Y_norm = array(rnorm(m * m * T_len, mean = 3, sd = 1), dim = c(m, m, T_len))
for (t in 1:T_len) diag(Y_norm[,,t]) = NA

X_norm = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X_norm[,,t] = Y_norm[,,t-1]
X_norm[is.na(X_norm)] = 0

fit_norm = sir(
	Y = Y_norm, W = W, X = X_norm, Z = Z,
	family = "normal", method = "ALS", calc_se = TRUE
)
summary(fit_norm)
```

### Binomial (binary outcomes)

For binary relational outcomes (e.g., presence or absence of a diplomatic tie, conflict onset), the Binomial family uses a logit link. The influence parameters now operate on the log-odds scale, so a positive $\alpha$ on a covariate indicates that the corresponding relationship characteristic increases the log-odds of influence.

```{r binomial}
Y_bin = array(rbinom(m * m * T_len, 1, 0.3), dim = c(m, m, T_len))
for (t in 1:T_len) diag(Y_bin[,,t]) = NA

X_bin = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X_bin[,,t] = Y_bin[,,t-1]
X_bin[is.na(X_bin)] = 0

fit_bin = sir(
	Y = Y_bin, W = W, X = X_bin, Z = Z,
	family = "binomial", method = "ALS", calc_se = TRUE
)
summary(fit_bin)
```

## Symmetric (undirected) networks

For undirected networks where $y_{i,j} = y_{j,i}$, setting `symmetric = TRUE` symmetrizes $\mathbf{Y}$, fits only the upper triangle, and constrains $\mathbf{B} = \mathbf{I}$. The sender/receiver distinction is meaningless for undirected data, so the model estimates only a single set of influence parameters. This is appropriate for networks such as undirected trade, co-sponsorship, or alliance formation.

```{r symmetric}
Y_sym = array(0, dim = c(m, m, T_len))
for (t in 1:T_len) {
	Y_t = matrix(rpois(m * m, 2), m, m)
	Y_sym[,,t] = (Y_t + t(Y_t)) / 2
	diag(Y_sym[,,t]) = NA
}
X_sym = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) {
	X_sym[,,t] = Y_sym[,,t-1]
	X_sym[,,t][is.na(X_sym[,,t])] = 0
}

fit_sym = sir(
	Y = Y_sym, W = W, X = X_sym, Z = Z,
	family = "poisson", symmetric = TRUE, calc_se = TRUE
)
fit_sym
```

## Bipartite networks

When the sender and receiver sets differ (rectangular $\mathbf{Y}$), bipartite structure is detected automatically and `fix_receiver = TRUE` is enforced. This is the appropriate specification for networks where the row and column actors represent distinct populations (e.g., countries sending aid to NGOs, or legislators co-sponsoring bills introduced by other legislators). For square arrays where senders and receivers are nonetheless distinct populations, set `bipartite = TRUE` explicitly.

```{r bipartite}
n1 = 10; n2 = 15
Y_bp = array(rpois(n1 * n2 * T_len, 2), dim = c(n1, n2, T_len))
W_bp = array(rnorm(n1 * n1 * p), dim = c(n1, n1, p))
X_bp = array(0, dim = c(n1, n2, T_len))
for (t in 2:T_len) X_bp[,,t] = log(Y_bp[,,t-1] + 1)
Z_bp = array(rnorm(n1 * n2 * 1 * T_len), dim = c(n1, n2, 1, T_len))

fit_bp = sir(
	Y = Y_bp, W = W_bp, X = X_bp, Z = Z_bp,
	family = "poisson", calc_se = TRUE
)
fit_bp
```

## Dynamic influence covariates

In many empirical settings, the factors that mediate influence change over time. Alliance structures evolve, trade relationships shift, and geographic proximity may be modulated by time-varying factors such as the presence of peacekeeping missions. When influence covariates change over time, supply `W` as a 4D array ($m \times m \times p \times T$). The influence parameters $\boldsymbol{\alpha}$ and $\boldsymbol{\beta}$ are still estimated as time-invariant quantities, but the influence matrices $\mathbf{A}_t$ and $\mathbf{B}_t$ now vary with $t$ because the underlying covariates do.

```{r dynamic-w}
Y_dyn = array(rpois(m * m * T_len, 2), dim = c(m, m, T_len))
for (t in 1:T_len) diag(Y_dyn[,,t]) = NA

X_dyn = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X_dyn[,,t] = log(Y_dyn[,,t-1] + 1)
X_dyn[is.na(X_dyn)] = 0

# 4D W: influence covariates that change over time
W_dyn = array(rnorm(m * m * p * T_len), dim = c(m, m, p, T_len))

fit_dyn = sir(
	Y = Y_dyn, W = W_dyn, X = X_dyn,
	family = "poisson", fix_receiver = TRUE,
	calc_se = FALSE, max_iter = 10
)

# A is now a 3D array (m x m x T)
dim(fit_dyn$A)
fit_dyn$dynamic_W
```

## Data preparation

### Converting edge lists to arrays

Network data often arrives as edge lists (long-format data frames with sender, receiver, time, and variable columns). The `cast_array()` function converts this format to the multidimensional arrays required by `sir()`.

```{r cast-array}
edge_list = expand.grid(
	i = paste0("n", 1:5),
	j = paste0("n", 1:5),
	t = 1:3
)
edge_list = edge_list[edge_list$i != edge_list$j, ]
edge_list$conflict = rpois(nrow(edge_list), lambda = 2)

Y_from_el = cast_array(edge_list, var = "conflict")
dim(Y_from_el)
dimnames(Y_from_el)[[1]]
dimnames(Y_from_el)[[3]]
```

### Constructing relational covariates

The `rel_covar()` function constructs relational covariate arrays from a base dyadic variable, creating main ($z_{ij}$), reciprocal ($z_{ji}$), and transitive ($\sum_k s_{ik} s_{kj}$ where $\mathbf{S}$ is the symmetrized network) effects. These derived quantities capture common relational mechanisms (reciprocity, transitivity) and can be included directly in the exogenous covariate array $\mathbf{Z}$.

```{r rel-covar}
# base trade array (m x m x T)
trade = array(abs(rnorm(m * m * T_len)), dim = c(m, m, T_len))

# create all three relational effects
Z_trade = rel_covar(trade, "trade")
dim(Z_trade)
dimnames(Z_trade)[[3]]
```

### Simulating SIR data

The `sim_sir()` function generates synthetic network data from the SIR data-generating process. This is useful for simulation studies, power analysis, and verifying that the estimation procedure recovers known parameter values under controlled conditions.

```{r sim-sir}
dat = sim_sir(m = 10, T_len = 8, p = 2, q = 1,
	family = "poisson", seed = 42)

str(dat[c("Y", "W", "alpha", "beta", "theta")])
```
