---
title: "Inference and model comparison"
author: "Shahryar Minhas & Peter Hoff"
description: "Choose classical, robust, cluster, bootstrap, jackknife, and predictive checks for SIR fits."
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Inference and model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = TRUE
)
library(sir)
```

Inference for bilinear models takes more care than for an ordinary GLM. Network data violate the independent-observations assumption behind naive Wald standard errors --- ties that share an actor are correlated --- so this vignette is mostly about getting *honest intervals* in the face of that dependence. (A separate risk, weak identification of the influence term, shows up as an ill-conditioned Hessian and is flagged by the `se_reliable` field, covered below.) It walks through classical and robust variance estimates, confidence intervals, bootstrap checks, and information-criterion model comparison. For basic fitting see `vignette("sir_overview")`.

### Getting standard errors, in one paragraph

Standard errors are computed by default (`calc_se = TRUE`), so you do not have to ask for them. **`summary(fit)` prints them, and `confint(fit)`, `vcov(fit)`, and `tidy(fit)` report cluster-robust intervals by default.** Cluster-robust is the right default for relational data: because ties that share an actor are correlated, classical (Hessian-based) errors that assume independent dyads can badly overstate significance. The coefficient table in `summary()`/`print()` shows the *classical* SE for a quick look (and labels it as such), but the interval you should report is the cluster-robust one from `confint()`. If the analytic standard errors cannot be formed --- a singular or ill-conditioned Hessian, or a full-bilinear bipartite fit, which has no closed-form covariance --- `sir()` **automatically falls back to the delete-one-actor jackknife**, and all four accessors use it transparently; the fit then carries `fit$se_source == "jackknife"` and says so in its printout. Everything below is the *why* behind these defaults, and how to cross-check them.

| Tool | What it is useful for | Main limitation |
|:-----|:----------------------|:----------------|
| Classical SE | Fast model-based Wald intervals; what `summary()`/`print()` show | Assumes independent dyad-period scores and a stable Hessian; can overstate significance |
| Cluster-robust SE (default) | Shared-actor dyadic dependence (each cell scored onto both actors); what `confint()`/`vcov()`/`tidy()` report | Still uses the Hessian as bread, so weak identification remains a problem |
| HC0 robust SE | Heteroskedasticity or overdispersion | Does not address shared-actor network dependence |
| Parametric bootstrap | Resimulates from the fitted model; confirms the model-based SEs | Assumes the model is correct, so it does not capture shared-actor dependence |
| Dyad jackknife | The automatic fallback when no analytic covariance exists (full-bilinear bipartite, ill-conditioned Hessian); also a conservative robustness check | Wider than the cluster interval; with few actors, deleting one is a large perturbation |

## Setup

We simulate a correctly specified Poisson model with `sim_sir()` so the standard errors are well behaved and we can focus on the mechanics.

```{r data}
set.seed(7)
dat <- sim_sir(m = 14, T_len = 80, p = 2, q = 2, family = "poisson", seed = 7)
fit <- sir(dat$Y, W = dat$W, X = dat$X, Z = dat$Z,
           family = "poisson", calc_se = TRUE, seed = 1)
```

The estimator recovers these known parameters (`vignette("sir_overview")` shows the recovery check); here we take the fit as given and focus on putting honest *intervals* around it. In the tables below, the coefficient labels tag each parameter's role --- `(Z)` a direct covariate effect, `(alphaW)` a sender-influence weight, `(betaW)` a receiver-influence weight --- and the first sender weight is fixed at 1 for identifiability, so there is no `(alphaW) W1` row.

## Classical (Hessian-based) standard errors

`sir(calc_se = TRUE)` computes classical standard errors from the inverse observed-information matrix. `summary()` prints them for a quick look, along with a note that they are Hessian-based and can overstate significance under dyadic dependence:

```{r summary, message = TRUE}
summary(fit)
```

The closing **Influence Matrices** block summarizes the fitted $m \times m$ sender and receiver matrices $A$ and $B$ (each a weighted sum of the `W` slices), so their entries live on a wider scale than the coefficient table above; `vignette("sir_overview")` covers how to read them.

For reporting, the default accessors use cluster-robust uncertainty when the fit supports it; request the classical SEs explicitly when you need a Hessian-only comparison:

```{r classical}
round(sqrt(diag(vcov(fit, type = "classical"))), 4)   # the SEs reported by summary()
confint(fit, se.type = "classical")                   # Hessian-only Wald intervals
```

## Robust (sandwich) standard errors

The HC0 sandwich estimator $H^{-1} S H^{-1}$ (with $S$ the empirical score covariance) corrects for heteroskedasticity or overdispersion. Our simulation is a correctly specified Poisson, so there is nothing to correct and the robust SEs match the classical ones almost exactly --- a useful confirmation that the model is well specified:

```{r robust}
se_classical <- sqrt(diag(vcov(fit, type = "classical")))
se_robust    <- sqrt(diag(vcov(fit, type = "robust")))

data.frame(
	term      = names(coef(fit)),
	classical = round(unname(se_classical), 4),
	robust    = round(unname(se_robust), 4),
	ratio     = round(unname(se_robust / se_classical), 2),
	row.names = NULL
)
```

To see HC0 do real work, refit on overdispersed counts --- here redrawn from a negative binomial, whose variance far exceeds its mean. Now the robust SEs come out about three times the classical ones, correctly widening for the extra dispersion the Poisson likelihood would otherwise ignore:

```{r robust-overdispersed}
set.seed(99)
Y_od <- dat$Y
Y_od[] <- rnbinom(length(Y_od), mu = pmax(dat$Y, 0.1), size = 0.5)   # heavier-than-Poisson counts
fit_od <- sir(Y_od, W = dat$W, X = dat$X, Z = dat$Z,
              family = "poisson", calc_se = TRUE, seed = 1)

data.frame(
	term = names(coef(fit_od)),
	robust_over_classical = round(sqrt(diag(vcov(fit_od, type = "robust"))) /
	                              sqrt(diag(vcov(fit_od, type = "classical"))), 2),
	row.names = NULL
)
```

HC0 handles dispersion and heteroskedasticity, but neither HC0 form models the shared-actor dependence that the **cluster-robust** estimator addresses next.

## Cluster-robust standard errors

Every tie involving an actor is correlated, so `vcov()` clusters on the **actors**: each dyad-period score is assigned to both of its endpoint actors and summed within actor. (Mechanically, an HC1 small-sample factor and a $t(G - 1)$ reference, with $G$ the number of actors, complete the sandwich.) This is the default reporting interval for supported static directed and symmetric fits when the Hessian bread is stable; report it in place of classical Hessian SEs:

```{r cluster}
se_cluster <- sqrt(diag(vcov(fit)))

data.frame(
	term      = names(coef(fit)),
	classical = round(unname(se_classical), 4),
	cluster   = round(unname(se_cluster), 4),
	ratio     = round(unname(se_cluster / se_classical), 2),
	row.names = NULL
)
```

Here the cluster-robust SEs are 1.0--1.9 times the classical ones: because they stop assuming independent dyad-times, they are the more conservative --- and the ones to report. `confint(fit)` returns the matching Wald intervals:

```{r cluster-confint}
confint(fit)
```

On this simulation the gap is modest because the periods are nearly independent. On real relational data it is not. Refitting on the bundled `icews` inter-state conflict network, the cluster SEs come out many times the classical ones:

```{r cluster-icews}
data(icews)
ic <- 1:10; ip <- 1:24
ifit <- sir(icews$Y[ic, ic, ip, drop = FALSE], W = icews$W[ic, ic, , drop = FALSE],
            X = icews$X[ic, ic, ip, drop = FALSE], Z = icews$Z[ic, ic, , ip, drop = FALSE],
            family = "poisson", seed = 1, max_iter = 20)

data.frame(
	term = names(coef(ifit)),
	cluster_over_classical = round(sqrt(diag(vcov(ifit, type = "cluster"))) /
	                               sqrt(diag(vcov(ifit, type = "classical"))), 1),
	row.names = NULL
)
```

Real conflict data carry exactly the shared-actor dependence the classical SEs assume away, so the classical intervals would badly overstate significance --- the cluster SEs are an order of magnitude larger. This is why the cluster interval is the default and the reported one.

## Bootstrap inference

`boot_sir()` refits the model on resampled data and forms distribution-free intervals that do not assume the Wald normal shape. Two are worth running: the **parametric** bootstrap resimulates outcomes from the fitted model (a model-based cross-check), and the **dyad jackknife** deletes each actor in turn (a conservative robustness check, and the fallback when analytic SEs are unavailable).

```{r boot}
br_par  <- boot_sir(fit, R = 200, type = "parametric", seed = 123, trace = FALSE)
br_dyad <- boot_sir(fit, type = "dyad", seed = 123, trace = FALSE)

data.frame(
	method = c("Parametric bootstrap", "Dyad jackknife"),
	valid_refits = c(br_par$n_valid, br_dyad$n_valid),
	total_refits = c(br_par$n_total, br_dyad$n_total),
	interval_type = c(br_par$interval, br_dyad$interval)
)
```

Each replicate refits the full model and non-converged ones are dropped, so check the valid-replicate count. The **parametric** bootstrap intervals track the classical ones and every one excludes zero --- a distribution-free confirmation that the estimates are well separated from zero:

```{r confint-par}
confint(fit, boot = br_par)
```

The **dyad jackknife** is deliberately conservative: deleting a whole actor is a large perturbation (especially with only `m = 14` actors), so its intervals run several times wider, most of all for the bilinear influence weights. Read it as a worst-case bound, not a competing precise estimate:

```{r confint-dyad}
confint(fit, boot = br_dyad)
```

For reporting, place the intervals side by side --- the classical (which ignores dyadic dependence), the cluster-robust default, and the conservative dyad jackknife:

```{r reporting-table}
fmt <- function(ci) sprintf("[%.3f, %.3f]", ci[, 1], ci[, 2])

data.frame(
	term           = names(coef(fit)),
	estimate       = round(unname(coef(fit)), 3),
	classical      = fmt(confint(fit, se.type = "classical")),
	cluster        = fmt(confint(fit)),
	dyad_jackknife = fmt(confint(fit, boot = br_dyad)),
	row.names = NULL
)
```

The estimate is shared across the columns (one fit); what changes is the interval. Classical is the narrowest (it assumes independent dyads), the cluster default widens it for shared-actor dependence, and the jackknife is the conservative outer bound. Report the cluster interval and use the jackknife as a robustness check. Use `R = 500`--`1000` for publication-quality bootstrap intervals.

## Choosing an approach

The choice reduces to three rules:

* **Report cluster-robust intervals** (`vcov()` / `confint()`, the default) for directed, symmetric, and dynamic (4D) `W` fits with a stable Hessian (`se_reliable = TRUE`).
* **Let the jackknife fallback handle the hard cases.** When no analytic covariance exists --- full-bilinear bipartite fits, or an ill-conditioned Hessian --- `sir()` already attaches the delete-one-actor jackknife automatically, so `confint()`/`vcov()`/`tidy()` keep working and `fit$se_source` reads `"jackknife"`; you do not need to call `boot_sir()` yourself. (You still can, e.g. `boot_sir(type = "dyad")` then `confint(fit, boot = bs)`, if you want to inspect the resampling.)
* **Do not** report dyad-independent classical $p$-values as the headline result for influence coefficients; pair the cluster interval with a response-scale scenario (below).

### How to tell which standard errors you have

`fit$se_source` is `"jackknife"` when the automatic fallback fired, and absent (NULL) otherwise --- meaning analytic SEs are available and `confint()`/`vcov()`/`tidy()` report the cluster-robust sandwich (including for dynamic, 4D `W`). The `summary(fit)` and `print(fit)` printouts also label the standard errors they show (classical in the table, jackknife when the fallback fired), so you are never left guessing which kind you are looking at.

## Fixed-receiver model and model comparison

Setting `fix_receiver = TRUE` constrains $B = I$ and estimates only sender-side influence. It is faster and better-conditioned, and serves as a constrained baseline. Because our simulation has a genuine receiver-side channel that this constraint drops, we expect the full model to fit better --- so the useful question is not what the constrained model's influence coefficients are, but whether dropping the receiver side costs fit. We answer that with information criteria rather than by reading the constrained coefficients:

```{r fix-receiver}
fit_fr <- sir(dat$Y, W = dat$W, X = dat$X, Z = dat$Z,
              family = "poisson", fix_receiver = TRUE, calc_se = TRUE, seed = 1)
```

Compare specifications with fit criteria and a common response-scale scenario. Because the two models differ in how the *lagged network* propagates, we shift the lagged signal `X` up by one standard deviation and read how each model's predicted mean responds:

```{r compare}
scenario_change <- function(model) {
	Xcf <- dat$X + stats::sd(dat$X, na.rm = TRUE)
	mu0 <- predict(model)
	mu1 <- predict(model, newdata = list(W = dat$W, X = Xcf, Z = dat$Z))
	mean(mu1 - mu0, na.rm = TRUE)
}

data.frame(
	model = c("full bilinear", "fix_receiver"),
	logLik = round(c(as.numeric(logLik(fit)), as.numeric(logLik(fit_fr))), 2),
	AIC   = round(c(AIC(fit), AIC(fit_fr)), 2),
	BIC   = round(c(BIC(fit), BIC(fit_fr)), 2),
	se_reliable = c(fit$se_reliable, fit_fr$se_reliable),
	mean_response_to_X = round(c(scenario_change(fit), scenario_change(fit_fr)), 3)
)
```

The `se_reliable` column flags whether the Hessian was well-conditioned enough to trust the analytic SEs (both fits here are reliable). Lower AIC/BIC is better, and the full bilinear model wins decisively --- confirming the receiver-side channel carries real signal. The scenario makes the difference concrete: shifting the lagged network moves the full model's predicted mean noticeably, but barely moves the fixed-receiver model's. Forcing $B = I$ leaves a sender map (the `A` matrix) that is both weaker in spread and nearly self-cancelling, so a uniform shift in the lagged network produces almost no net change in its predicted mean, whereas the full model's stronger, coherent influence structure responds. The much larger drop in log-likelihood for the constrained model confirms the receiver side carries signal it cannot recover. `vignette("sir_overview")` covers `predict(newdata = )` scenarios in full.

## Out-of-sample evaluation

In-sample criteria (AIC/BIC) reward complexity; out-of-sample accuracy is the more demanding test. `forecast()` produces iterated plug-in forecasts, and `cv_sir()` runs rolling-origin (expanding-window) cross-validation, scoring each one-step forecast against the held-out actuals alongside a last-value naive baseline. `cv_sir()` conditions on the future `Z` and dynamic-`W` slices stored on the fitted object, so treat its scores as a conditional validation target: the future covariates must be known, pre-specified, or generated from a scenario available at the forecast origin.

```{r forecast-cv}
# conditional one-step forecast: fit through T-1, then pass a Z_T slice that is
# known, fixed, or scenario-generated at the forecast origin. In this simulated
# example, the held-out Z_T is used to show the API.
T_hold <- dim(dat$Y)[3] - 1
fit_train <- sir(dat$Y[, , 1:T_hold, drop = FALSE],
                 W = dat$W,
                 X = dat$X[, , 1:T_hold, drop = FALSE],
                 Z = dat$Z[, , , 1:T_hold, drop = FALSE],
                 family = "poisson", calc_se = FALSE, seed = 1)
Z_hold <- dat$Z[, , , T_hold + 1, drop = FALSE]
fc <- forecast(fit_train, h = 1, Z_future = Z_hold)
model_score <- score_sir(dat$Y[, , T_hold + 1, drop = FALSE], fc, "poisson")
naive_score <- score_sir(
	dat$Y[, , T_hold + 1, drop = FALSE],
	array(dat$Y[, , T_hold], dim = dim(fc)),
	"poisson"
)
data.frame(
	metric = names(model_score),
	model = round(unname(model_score), 3),
	naive = round(unname(naive_score), 3),
	improvement = round(unname(naive_score - model_score), 3),
	row.names = NULL
)

# rolling-origin conditional cross-validation: initial is the minimum training
# length and origins are the cut-points o (each refits on 1..o and scores o + 1)
cv <- cv_sir(fit, initial = 60, origins = c(60, 70, 79))
data.frame(
	metric = c("rmse", "mae", "deviance"),
	model = round(unname(cv$aggregate[c("rmse", "mae", "deviance")]), 3),
	naive = round(unname(cv$aggregate[paste0("naive_", c("rmse", "mae", "deviance"))]), 3),
	improvement = round(unname(cv$aggregate[paste0("naive_", c("rmse", "mae", "deviance"))] -
	                           cv$aggregate[c("rmse", "mae", "deviance")]), 3),
	effective_origins = unname(cv$eff_n[c("rmse", "mae", "deviance")]),
	row.names = NULL
)
```

Lower RMSE/MAE/deviance scores are better. The model improves all three over the naive last-value baseline; the deviance gap is the largest because Poisson deviance penalizes the naive predictor heavily when its carried-forward counts miss, while RMSE and MAE are gentler. This example uses only three origins, so it demonstrates the workflow rather than providing a final validation study; with real applications, use more origins or a substantively chosen rolling schedule.

`score_sir()` exposes the same family-appropriate scores (RMSE / MAE plus the Poisson deviance) for any observed/predicted pair, so you can assemble custom backtests.
