---
title: "Distribution families, network types, and data preparation"
author: "Shahryar Minhas & Peter Hoff"
description: "Use Normal and Binomial outcomes, symmetric and bipartite networks, dynamic influence covariates, and array builders."
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Distribution families, network types, and data preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = TRUE
)
library(sir)
```

The SIR framework handles a range of outcome types, network structures, and
covariate configurations beyond the Poisson directed-network case in the
overview. This vignette covers the distribution families, symmetric and bipartite
networks, dynamic (time-varying) influence covariates, and the data-preparation
utilities. For basic fitting see `vignette("sir_overview")`.

## Distribution families

The distribution-family examples below use `sim_sir()`, so each fits data with a
*known* influence signal and the coefficients are interpretable. The later
sections (symmetric, bipartite, dynamic) use small illustrative networks to show
the *call mechanics* of each structure; there the simulated outcome is plain
noise, so the fitted coefficients are not meaningful.

### Normal (continuous outcomes)

For continuous relational data (trade volumes, sentiment scores) the Normal
family uses an identity link. The influence parameters are interpreted exactly as
in the Poisson case: $\alpha$ and $\beta$ identify which covariates account for
sender and receiver influence, with effects entering the linear predictor.

```{r normal}
dn  <- sim_sir(m = 14, T_len = 80, p = 2, q = 2, family = "normal", seed = 11)
fit_norm <- sir(dn$Y, W = dn$W, X = dn$X, Z = dn$Z,
                family = "normal", calc_se = FALSE, seed = 1)
target_norm <- c(dn$theta, dn$alpha[-1], dn$beta)
data.frame(
	term = names(coef(fit_norm)),
	estimate = round(unname(coef(fit_norm)), 3),
	target = round(target_norm, 3),
	abs_error = round(abs(unname(coef(fit_norm)) - target_norm), 3),
	row.names = NULL
)
```

The synthetic data were generated from a known Normal SIR process, so the
coefficient signs and magnitudes are meaningful for this example. We show
recovery against the generating values rather than a star table because this
section is about family behavior, not inference.

On the response scale, the Normal family returns fitted means directly:

```{r normal-response}
mu_norm <- predict(fit_norm)
round(quantile(mu_norm[!is.na(mu_norm)], c(0.05, 0.5, 0.95)), 3)
```

### Binomial (binary outcomes)

For binary relational outcomes (presence/absence of a tie, conflict onset) the
Binomial family uses a logit link. The full bilinear contribution
$(A X_t B^\top)_{ij}$ enters the log-odds, so an individual $\alpha$ or $\beta$
coefficient is not a standalone odds-ratio term; its effect depends on `X`, the
other influence side, and the covariate scale.

```{r binomial}
db  <- sim_sir(m = 16, T_len = 120, p = 2, q = 1, family = "binomial", seed = 12)
fit_bin <- sir(db$Y, W = db$W, X = db$X, Z = db$Z,
               family = "binomial", calc_se = FALSE, seed = 1)
target_bin <- c(db$theta, db$alpha[-1], db$beta)
data.frame(
	term = names(coef(fit_bin)),
	estimate = round(unname(coef(fit_bin)), 3),
	target = round(target_bin, 3),
	abs_error = round(abs(unname(coef(fit_bin)) - target_bin), 3),
	row.names = NULL
)
```

Again, these are simulated data with a known generating process, so recovery and
response-scale probabilities are the useful outputs.

`predict()` returns fitted tie probabilities by default. Use
`predict(type = "link")` and then `plogis()` only when you explicitly request
the log-odds scale:

```{r binomial-prob}
mu_bin <- predict(fit_bin)                 # fitted tie probabilities
round(quantile(mu_bin[!is.na(mu_bin)], c(0.05, 0.5, 0.95)), 3)
```

## Symmetric (undirected) networks

For undirected networks where $y_{i,j} = y_{j,i}$, set `symmetric = TRUE`. The
model fits the upper triangle and constrains $B = I$. This is an upper-triangle,
sender-side representation of undirected data, not a fully order-invariant
undirected bilinear model. For Poisson and Binomial outcomes, provide a symmetric
`Y` directly; averaging asymmetric discrete matrices can create invalid
non-integer or non-binary observations. The influence covariates in `W` must also
be symmetric and meaningful for unordered pairs. This suits undirected trade,
alliance, or co-sponsorship networks when that representation is substantively
acceptable.

```{r symmetric}
set.seed(1)
m <- 14; T_len <- 60; p <- 2
W <- array(0, dim = c(m, m, p))
for (k in seq_len(p)) {
	W_k <- matrix(rnorm(m * m), m, m)
	W[, , k] <- (W_k + t(W_k)) / 2
	diag(W[, , k]) <- 0
}

Y_sym <- array(0, dim = c(m, m, T_len))
for (t in 1:T_len) {
	Y_t <- matrix(0, m, m)
	upper <- upper.tri(Y_t)
	Y_t[upper] <- rpois(sum(upper), 2)
	Y_t <- Y_t + t(Y_t)
	diag(Y_t) <- NA
	Y_sym[, , t] <- Y_t
}
X_sym <- array(0, dim = c(m, m, T_len))
for (t in 2:T_len) {
	X_sym[, , t] <- log(Y_sym[, , t - 1] + 1)
	X_sym[, , t][is.na(X_sym[, , t])] <- 0
}

fit_sym <- sir(Y_sym, W = W, X = X_sym, family = "poisson",
               symmetric = TRUE, calc_se = FALSE, seed = 1)
data.frame(
	converged = fit_sym$convergence,
	symmetric = fit_sym$symmetric,
	fix_receiver = fit_sym$fix_receiver
)
```

This chunk demonstrates the API and structural checks. The outcome was generated
as simple noise, so the fitted coefficient values are not substantively
interpretable.

> These illustrative chunks build `X` as a plain `log(Y_prev + 1)` for brevity.
> When you intend to forecast, build `X` the way `sim_sir()` and `forecast()`
> assume --- divide the lag by `max(m - 1, 1)` (the number of partners summed
> over) --- or pass a matching `infl_scale` to `forecast()`.

## Bipartite networks

When senders and receivers are **distinct populations** (a rectangular $Y$, e.g.
countries $\to$ NGOs, or legislators co-sponsoring others' bills), bipartite
structure is detected automatically and `fix_receiver = TRUE` is enforced.

The detail that trips people up: **influence flows on the sender side, so `W`
describes the senders.** For an $n_1 \times n_2$ outcome, `W` must be
$n_1 \times n_1 \times p$ (sender-by-sender), *not* keyed to the receivers. `X`
matches `Y` at $n_1 \times n_2$.

```{r bipartite}
set.seed(2)
n1 <- 10; n2 <- 15; T_len <- 40; p <- 2
Y_bp <- array(rpois(n1 * n2 * T_len, 2), dim = c(n1, n2, T_len))
W_bp <- array(rnorm(n1 * n1 * p), dim = c(n1, n1, p))   # sender-by-sender
X_bp <- array(0, dim = c(n1, n2, T_len))
for (t in 2:T_len) X_bp[, , t] <- log(Y_bp[, , t - 1] + 1)
Z_bp <- array(rnorm(n1 * n2 * 1 * T_len), dim = c(n1, n2, 1, T_len))

fit_bp <- sir(Y_bp, W = W_bp, X = X_bp, Z = Z_bp,
              family = "poisson", fix_receiver = TRUE, calc_se = FALSE, seed = 1)
data.frame(
	converged = fit_bp$convergence,
	bipartite = fit_bp$bipartite,
	fix_receiver = fit_bp$fix_receiver,
	senders = fit_bp$n1,
	receivers = fit_bp$n2
)
```

Again, this is an API demonstration: the simulated outcome is plain noise, so
only the structural output should be read.

For a *square* array whose rows and columns are nonetheless distinct populations,
set `bipartite = TRUE` explicitly so the receiver side is not estimated.

### Full bilinear bipartite (with `W_recv`)

The default above collapses `B` to the identity. To estimate the full bilinear
model $A X B'$ for two-mode data --- with a *separate* receiver-side influence
structure --- supply `W_recv`, an $n_2 \times n_2 \times p_2$ array. The sender
influence `A` is then built from `W` and the receiver influence `B` from
`W_recv`, fit by alternating GLM with `alpha_1 = 1` pinning the scale.

Here we simulate a two-mode network *with* genuine bilinear structure so the
fit has something to recover (the earlier `Y_bp` was pure noise, useful only to
show the API). True sender weights are `alpha = (1, 0.6)` and receiver weights
`beta = (0.8, -0.5)`:

```{r bipartite-full}
set.seed(204)
n1 <- 10; n2 <- 6; Tn <- 30
Wf  <- array(rnorm(n1 * n1 * 2), dim = c(n1, n1, 2))   # sender-by-sender
Wr  <- array(rnorm(n2 * n2 * 2), dim = c(n2, n2, 2))   # receiver-by-receiver
Xf  <- array(rnorm(n1 * n2 * Tn) / sqrt(n2), dim = c(n1, n2, Tn))
A_t <- Wf[, , 1] + 0.6 * Wf[, , 2]
B_t <- 0.8 * Wr[, , 1] - 0.5 * Wr[, , 2]
Yf  <- array(0, dim = c(n1, n2, Tn))
for (t in 1:Tn) Yf[, , t] <- A_t %*% Xf[, , t] %*% t(B_t) +
    matrix(rnorm(n1 * n2, sd = 0.3), n1, n2)

fit_full <- sir(Yf, W = Wf, X = Xf, W_recv = Wr,
                family = "normal", calc_se = FALSE, seed = 1)
data.frame(
	full_bilinear = isTRUE(fit_full$full_bilinear),
	converged = fit_full$convergence
)

data.frame(
	term = names(coef(fit_full)),
	estimate = round(unname(coef(fit_full)), 3),
	target = c(0.6, 0.8, -0.5),
	row.names = NULL
)
```

Analytic standard errors are not available on this path; use the delete-one-actor
dyad jackknife for inference:

```{r bipartite-full-boot}
bj <- boot_sir(fit_full, type = "dyad", seed = 1)
confint(bj)
```

## Dynamic influence covariates

Often the factors that mediate influence change over time --- alliances evolve,
trade shifts. Supply `W` as a 4D array ($m \times m \times p \times T$). The
coefficients $\boldsymbol{\alpha}$ and $\boldsymbol{\beta}$ are still estimated as
time-invariant, but the reconstructed influence matrices $A_t$ now vary with $t$
because the covariates do.

```{r dynamic-w}
set.seed(3)
m <- 14; T_len <- 40; p <- 2
Y_dyn <- array(rpois(m * m * T_len, 2), dim = c(m, m, T_len))
for (t in 1:T_len) diag(Y_dyn[, , t]) <- NA
X_dyn <- array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X_dyn[, , t] <- log(Y_dyn[, , t - 1] + 1)
X_dyn[is.na(X_dyn)] <- 0
W_dyn <- array(rnorm(m * m * p * T_len), dim = c(m, m, p, T_len))   # 4D, time-varying

fit_dyn <- sir(Y_dyn, W = W_dyn, X = X_dyn, family = "poisson",
               fix_receiver = TRUE, calc_se = FALSE, max_iter = 10, seed = 1)

dim(fit_dyn$A)        # A is now m x m x T
fit_dyn$dynamic_W
```

The coefficients are fixed across time, but the realized channels change with
`W_t`. Because this example uses simulated noise to demonstrate the API, the
right rendered output is a structural check rather than a ranked channel table:

```{r dynamic-w-edges}
dynamic_period_check <- function(period) {
	A_t <- fit_dyn$A[, , period]
	W_t <- W_dyn[, , , period]
	diag(A_t) <- NA
	data.frame(
		period = period,
		A_dimensions = paste(dim(A_t), collapse = " x "),
		finite_off_diagonal = sum(is.finite(A_t)),
		mean_abs_A = round(mean(abs(A_t), na.rm = TRUE), 3),
		mean_abs_W = round(mean(abs(W_t), na.rm = TRUE), 3),
		row.names = NULL
	)
}
rbind(
	dynamic_period_check(1),
	dynamic_period_check(T_len)
)
```

## Data preparation

### Converting edge lists to arrays

Network data often arrives as a long-format edge list (sender, receiver, time,
value). `cast_array()` converts it to the arrays `sir()` expects.

```{r cast-array}
set.seed(4)
edge_list <- expand.grid(i = paste0("n", 1:5), j = paste0("n", 1:5), t = 1:3)
edge_list <- edge_list[edge_list$i != edge_list$j, ]
edge_list$conflict <- rpois(nrow(edge_list), lambda = 2)

Y_from_el <- cast_array(edge_list, var = "conflict")
dim(Y_from_el)
dimnames(Y_from_el)[[1]]
```

### Constructing relational covariates

`rel_covar()` builds main ($z_{ij}$), reciprocal ($z_{ji}$), and transitive
($\sum_k s_{ik} s_{kj}$ on the symmetrized network) effects from a base dyadic
variable. These capture common relational mechanisms and drop straight into the
exogenous covariate array $Z$.

```{r rel-covar}
set.seed(5)
trade   <- array(abs(rnorm(8 * 8 * 4)), dim = c(8, 8, 4))   # base dyadic variable
Z_trade <- rel_covar(trade, "trade")
dim(Z_trade)
dimnames(Z_trade)[[3]]
```

### Simulating SIR data

`sim_sir()` (used throughout these vignettes) generates synthetic data from the
SIR process --- useful for simulation studies, power analysis, and verifying
parameter recovery under controlled conditions.

```{r sim-sir}
dat <- sim_sir(m = 10, T_len = 8, p = 2, q = 1, family = "poisson", seed = 42)
str(dat[c("Y", "W", "X", "Z", "alpha", "beta", "theta")], max.level = 1)
```
