---
title: "Getting started with sir"
author: "Shahryar Minhas & Peter Hoff"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with sir}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = FALSE
)
library(sir)
library(ggplot2)
```

## Why SIR?

When country $i$ initiates conflict with country $k$, does that make it more likely that country $j$ --- an ally of $i$ --- does the same? Standard regression treats each directed dyad as independent, so it cannot answer this question. The Social Influence Regression (SIR) model can: it regresses current network outcomes on the *entire* lagged network, decomposing who influences whom and estimating which observable covariates (alliances, proximity, shared group membership) drive those influence patterns.

The key equation is:

$$\mu_{i,j,t} = \boldsymbol{\theta}^\top \mathbf{z}_{i,j,t} + \sum_{k,\ell} x_{k,\ell,t} \, a_{i,k} \, b_{j,\ell}$$

The first term is standard regression on exogenous covariates. The second is the influence term: $a_{i,k}$ measures how predictive node $k$'s past sending behavior is of node $i$'s, and $b_{j,\ell}$ captures an analogous receiver-side channel. Rather than estimating every $a_{i,k}$ freely ($O(n^2)$ parameters), SIR parameterizes the influence matrices through covariates:

$$\mathbf{A} = \sum_{r=1}^{p} \alpha_r \mathbf{W}_r, \qquad \mathbf{B} = \sum_{r=1}^{p} \beta_r \mathbf{W}_r$$

This reduces the problem to estimating a handful of $\alpha$ and $\beta$ coefficients that tell you *which* covariates matter for influence and by how much. See `vignette("methodology")` for the full mathematical framework.

## Quick start: simulate, fit, recover

The fastest way to see SIR in action is to simulate data with known parameters and check that the model recovers them. We set up a small directed network of 15 nodes over 10 time periods with Poisson counts, two influence covariates (geographic proximity and shared group membership), and one exogenous covariate (dyadic distance).

```{r simulate}
set.seed(42)
m = 15; T_len = 10; p = 2

# influence covariates
W = array(0, dim = c(m, m, p))
geo = matrix(runif(m * m), m, m)
geo = (geo + t(geo)) / 2; diag(geo) = 0
W[,,1] = geo

groups = sample(1:3, m, replace = TRUE)
W[,,2] = (outer(groups, groups, "==")) * 1.0; diag(W[,,2]) = 0
dimnames(W) = list(paste0("n", 1:m), paste0("n", 1:m),
	c("proximity", "shared_group"))

# true parameters
alpha_true = c(1, 0.3)     # sender influence (alpha_1 = 1 fixed)
beta_true  = c(0.5, 0.2)   # receiver influence
theta_true = -0.1           # exogenous covariate effect

# build true influence matrices
A_true = alpha_true[1] * W[,,1] + alpha_true[2] * W[,,2]
B_true = beta_true[1] * W[,,1] + beta_true[2] * W[,,2]

# exogenous covariate: dyadic distance (time-invariant)
Z = array(0, dim = c(m, m, 1, T_len))
distance = matrix(rnorm(m * m), m, m)
distance = (distance + t(distance)) / 2; diag(distance) = NA
for (t in 1:T_len) Z[,,1,t] = distance
dimnames(Z)[[3]] = "distance"

# simulate Poisson network with influence
Y = array(NA, dim = c(m, m, T_len))
Y[,,1] = matrix(rpois(m * m, lambda = 2), m, m); diag(Y[,,1]) = NA
for (t in 2:T_len) {
	X_t = log(Y[,,t-1] + 1); X_t[is.na(X_t)] = 0
	eta = theta_true * Z[,,1,t] + A_true %*% X_t %*% t(B_true)
	Y[,,t] = matrix(rpois(m * m, pmin(exp(eta), 100)), m, m)
	diag(Y[,,t]) = NA
}

# X = log-transformed lagged Y
X = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X[,,t] = log(Y[,,t-1] + 1)
X[is.na(X)] = 0
```

Now fit the model:

```{r fit}
fit = sir(
	Y = Y, W = W, X = X, Z = Z,
	family = "poisson",
	method = "ALS",
	calc_se = TRUE
)
```

### Parameter recovery

The point of simulated data is verification. Let's compare the estimates to the truth:

```{r recovery}
true_tab = c(theta_true, alpha_true[-1], beta_true)
est_tab  = coef(fit)

recovery = data.frame(
	true      = round(true_tab, 3),
	estimated = round(est_tab, 3),
	se        = round(fit$summ$se, 3),
	covered   = abs(true_tab - est_tab) < 1.96 * fit$summ$se
)
recovery
```

The model recovers the parameters well: estimated values are close to the truth, and all true values fall within their 95% confidence bands.

### Model summary

```{r summary}
summary(fit)
```

The summary reports both classical (Hessian-based) and robust (sandwich) standard errors. For bilinear models the Hessian can be ill-conditioned; when the two sets of SEs diverge substantially, bootstrap inference via `boot_sir()` is recommended (see `vignette("sir_inference")`).

## Interpreting influence

The estimated $\alpha$ and $\beta$ coefficients tell you which covariates drive network influence:

- A positive $\alpha_r$ means that dyads with higher values on covariate $W_r$ exhibit stronger *sender-side* influence: if $w_{i,k}$ is large, then $k$'s past sending behavior is more predictive of $i$'s current sending.
- A positive $\beta_r$ means the same on the *receiver side*: if $w_{j,\ell}$ is large, then past targeting of $\ell$ predicts current targeting of $j$.

In our example, the positive coefficients on both `proximity` and `shared_group` mean that geographically close countries and countries in the same group tend to influence each other's conflict behavior.

The influence matrices $\mathbf{A}$ and $\mathbf{B}$ are the weighted sums of the $W$ covariates. Their off-diagonal entries summarize the full influence structure:

```{r influence}
A_hat = fit$A
A_offdiag = A_hat[row(A_hat) != col(A_hat)]
cat("Sender influence (A) — off-diagonal summary:\n")
summary(A_offdiag)

B_hat = fit$B
B_offdiag = B_hat[row(B_hat) != col(B_hat)]
cat("\nReceiver influence (B) — off-diagonal summary:\n")
summary(B_offdiag)
```

## Diagnostic plots

The `plot()` method provides several diagnostics. Panels 1--4 show heatmaps of the influence matrices and the distribution of their off-diagonal entries, which together indicate whether influence is concentrated among particular dyads or distributed broadly:

```{r plot-diagnostics, fig.width=8, fig.height=8}
plot(fit, which = 1:4, title = "Influence Structure")
```

The convergence trace (panel 5) confirms that ALS has stabilized:

```{r plot-convergence, fig.width=6, fig.height=4}
plot(fit, which = 5, combine = FALSE)
```

The coefficient plot (panel 6) visualizes parameter estimates with confidence intervals:

```{r plot-coefs, fig.width=6, fig.height=4}
plot(fit, which = 6, combine = FALSE)
```

## Prediction

`predict()` returns fitted values on either the link or response scale:

```{r predict}
pred = predict(fit, type = "response")

# compare predicted vs observed for a single time slice
obs = Y[,,5]
pred_t5 = pred[,,5]
mask = !is.na(obs)
cor(c(obs[mask]), c(pred_t5[mask]))
```

For out-of-sample prediction, pass new data via the `newdata` argument:

```{r predict-new, eval = FALSE}
predict(fit, newdata = list(Y = Y_new, W = W, X = X_new, Z = Z_new),
	type = "response")
```

## Data format

The package expects data as multidimensional arrays:

| Array | Dimensions | Description |
|:------|:-----------|:------------|
| `Y` | $m \times m \times T$ | Network outcomes (counts, continuous, or binary) |
| `X` | $m \times m \times T$ | Network state carrying influence (typically lagged `Y`) |
| `W` | $m \times m \times p$ | Influence covariates parameterizing $\mathbf{A}$ and $\mathbf{B}$ |
| `Z` | $m \times m \times q \times T$ | Exogenous dyadic covariates (optional) |

Long-format edge lists can be converted with `cast_array()`, and `rel_covar()` constructs relational covariates (main, reciprocal, transitive effects) from a base dyadic variable. See `vignette("sir_extensions")` for details.

## Next steps

- **[Methodology](methodology.html)**: Full mathematical framework, identifiability, estimation algorithms, and guidance on choosing a model configuration.
- **[Inference](sir_inference.html)**: Variance-covariance estimation, fixed-receiver models, bootstrap standard errors, and confidence intervals.
- **[Extensions](sir_extensions.html)**: Normal and binomial families, symmetric and bipartite networks, dynamic (time-varying) influence covariates, and data preparation utilities.
