---
title: "Inference and model comparison"
author: "Shahryar Minhas & Peter Hoff"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Inference and model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = FALSE
)
library(sir)
```

Inference for bilinear models requires more care than for standard GLMs. The bilinear structure in the SIR model can produce ill-conditioned Hessians, which in turn yield unreliable Wald-based standard errors. This vignette covers the inference tools available in the package: classical and robust variance-covariance estimation, fixed-receiver models that circumvent the identification problem, bootstrap standard errors, and confidence intervals. For basic model fitting see `vignette("sir_overview")`.

## Setup

We use the same simulated data as in the overview vignette.

```{r data}
set.seed(42)
m = 15; T_len = 10; p = 2

W = array(0, dim = c(m, m, p))
geo = matrix(runif(m * m), m, m); geo = (geo + t(geo)) / 2; diag(geo) = 0
W[,,1] = geo
groups = sample(1:3, m, replace = TRUE)
W[,,2] = outer(groups, groups, "==") * 1.0; diag(W[,,2]) = 0
dimnames(W) = list(paste0("n", 1:m), paste0("n", 1:m), c("proximity", "shared_group"))

Z = array(0, dim = c(m, m, 1, T_len))
dist_mat = matrix(rnorm(m * m), m, m); dist_mat = (dist_mat + t(dist_mat)) / 2
diag(dist_mat) = NA
for (t in 1:T_len) Z[,,1,t] = dist_mat
dimnames(Z)[[3]] = "distance"

alpha_true = c(1, 0.3); beta_true = c(0.5, 0.2); theta_true = -0.1
A_true = alpha_true[1] * W[,,1] + alpha_true[2] * W[,,2]
B_true = beta_true[1] * W[,,1] + beta_true[2] * W[,,2]

Y = array(NA, dim = c(m, m, T_len))
Y[,,1] = matrix(rpois(m * m, 2), m, m); diag(Y[,,1]) = NA
for (t in 2:T_len) {
	X_t = log(Y[,,t-1] + 1); X_t[is.na(X_t)] = 0
	eta = theta_true * Z[,,1,t] + A_true %*% X_t %*% t(B_true)
	lambda = pmin(exp(eta), 100)
	Y[,,t] = matrix(rpois(m * m, c(lambda)), m, m); diag(Y[,,t]) = NA
}

X = array(0, dim = c(m, m, T_len))
for (t in 2:T_len) X[,,t] = log(Y[,,t-1] + 1)
X[is.na(X)] = 0
```

```{r fit}
fit_als = sir(
	Y = Y, W = W, X = X, Z = Z,
	family = "poisson", method = "ALS", calc_se = TRUE
)
```

## Variance-covariance estimation

The package provides both classical (Hessian-based) and robust (sandwich) variance-covariance estimates. The classical estimate $H^{-1}$ is efficient when the model is correctly specified. The sandwich estimator $H^{-1} S H^{-1}$, where $S$ is the empirical score covariance, remains valid under misspecification. In practice, the two estimates often diverge for the bilinear parameters more than for the exogenous covariate coefficients, which can itself serve as a diagnostic for model adequacy.

```{r vcov}
V_classical = vcov(fit_als, type = "classical")
V_robust = vcov(fit_als, type = "robust")

# classical and robust standard errors
if (!is.null(V_classical)) sqrt(diag(V_classical))
if (!is.null(V_robust)) sqrt(diag(V_robust))
```

## Fixed-receiver model

Setting `fix_receiver = TRUE` constrains $\mathbf{B} = \mathbf{I}$ and estimates only sender-side influence. This eliminates the bilinear identification problem entirely and reduces the model to a standard GLM with well-conditioned standard errors. When the research question concerns sender-side influence alone, the fixed-receiver specification is a natural starting point. It also provides a useful baseline for comparison with the full bilinear model.

```{r fix-receiver}
fit_fix = sir(
	Y = Y, W = W, X = X, Z = Z,
	family = "poisson",
	method = "ALS",
	fix_receiver = TRUE,
	calc_se = TRUE
)

summary(fit_fix)
```

### Model comparison

Information criteria provide a straightforward basis for comparing nested specifications. The full bilinear model introduces additional receiver-side parameters; whether the improvement in fit justifies the added complexity is an empirical question.

```{r compare}
AIC(fit_als)
AIC(fit_fix)
BIC(fit_als)
BIC(fit_fix)
```

## Bootstrap inference

When the Hessian is ill-conditioned, Wald-based standard errors become unreliable. Bootstrap standard errors address this by constructing an empirical sampling distribution for the parameter estimates. The block bootstrap resamples entire time periods, preserving the temporal dependence structure within each period. A parametric bootstrap, which simulates new outcome arrays from the fitted model, is also available.

```{r boot, cache = TRUE}
boot_results = boot_sir(
	fit_als,
	R = 100,
	type = "block",
	seed = 123,
	trace = FALSE
)

boot_results
```

## Confidence intervals

Wald-based confidence intervals use the Hessian standard errors and assume approximate normality of the sampling distribution.

```{r confint}
confint(fit_als)
```

When bootstrap results are available, percentile intervals constructed from the empirical distribution of bootstrap replicates can be used instead. These are generally more reliable for bilinear parameters, particularly when the Hessian is ill-conditioned.

```{r confint-boot}
confint(fit_als, boot = boot_results)
```
