---
title: "Inference and model comparison"
author: "Shahryar Minhas & Peter Hoff"
description: "Choose classical, robust, cluster, bootstrap, jackknife, and predictive checks for SIR fits."
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Inference and model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = TRUE
)
library(sir)
```

Inference for bilinear models takes more care than for an ordinary GLM. The
bilinear influence term can produce an ill-conditioned Hessian, which makes naive
Wald standard errors unreliable. This vignette covers the tools the package
provides: classical and robust variance estimates, confidence intervals,
bootstrap and jackknife checks, and information-criterion model comparison. For
basic fitting see `vignette("sir_overview")`.

| Tool | What it is useful for | Main limitation |
|:-----|:----------------------|:----------------|
| Classical SE | Fast model-based Wald intervals | Assumes independent dyad-period scores and a stable Hessian |
| HC0 robust SE | Heteroskedasticity or overdispersion | Does not address shared-actor network dependence |
| Multiway cluster SE | Sender, receiver, and time dependence in the score | Still uses the Hessian as bread, so weak identification remains a problem |
| Block bootstrap | Time-period resampling | Preserves within-period dependence but not actor dependence |
| Dyad jackknife | Delete-one-actor sensitivity and actor dependence checks | Normal jackknife intervals, not bootstrap percentile intervals |

## Setup

We simulate a correctly specified Poisson model with `sim_sir()` so the standard
errors are well behaved and we can focus on the mechanics.

```{r data}
set.seed(7)
dat <- sim_sir(m = 14, T_len = 80, p = 2, q = 2, family = "poisson", seed = 7)
fit <- sir(dat$Y, W = dat$W, X = dat$X, Z = dat$Z,
           family = "poisson", calc_se = TRUE, seed = 1)
```

## Classical (Hessian-based) standard errors

`sir(calc_se = TRUE)` computes classical standard errors from the inverse
observed-information matrix. These are what `summary()` prints for quick model
inspection. For reporting, the default accessors now use cluster-robust
uncertainty when the fit supports it; request classical intervals explicitly
when you need a Hessian-only comparison:

```{r classical}
sqrt(diag(vcov(fit, type = "classical")))   # the SEs reported by summary()
confint(fit, se.type = "classical")         # Hessian-only Wald intervals
```

## Robust (sandwich) standard errors

The HC0 sandwich estimator $H^{-1} S H^{-1}$ (with $S$ the empirical score
covariance) is useful for heteroskedasticity or overdispersion when dyad-period
score contributions are otherwise independent and the Hessian bread is stable.
Compare it against the classical SEs:

```{r robust}
se_classical <- sqrt(diag(vcov(fit, type = "classical")))
se_robust    <- sqrt(diag(vcov(fit, type = "robust")))

data.frame(
	term      = names(coef(fit)),
	classical = round(unname(se_classical), 4),
	robust    = round(unname(se_robust), 4),
	ratio     = round(unname(se_robust / se_classical), 2),
	row.names = NULL
)

confint(fit, se.type = "robust")
```

For well-specified data the HC0 robust SEs barely move (ratio near 1): HC0
corrects for overdispersion, not for the dyadic dependence that dominates
relational data. That dependence is what the **cluster-robust** estimator
addresses.

## Multiway cluster-robust standard errors

The $(i, j)$ and $(j, i)$ dyads share both actors, and every tie involving an
actor is correlated. `vcov()` clusters the score contributions
on the sender, receiver, and time margins (`"twoway"` remains an alias for this
multiway estimator). For supported static directed-network fits, report this
interval in place of classical Hessian SEs when the Hessian bread is stable:

```{r cluster}
se_cluster <- sqrt(diag(vcov(fit)))

data.frame(
	term      = names(coef(fit)),
	classical = round(unname(se_classical), 4),
	cluster   = round(unname(se_cluster), 4),
	ratio     = round(unname(se_cluster / se_classical), 2),
	row.names = NULL
)
```

The cluster-robust SEs can move in either direction in this clean simulation. In
dependent real data they are often more conservative because they stop assuming
independent dyad-times. `confint(fit)` returns the matching Wald intervals:

```{r cluster-confint}
confint(fit)
```

## Bootstrap inference

`boot_sir()` refits the model on resampled or reduced data. The **block**
bootstrap resamples whole time periods (preserving within-period dependence);
the **dyad jackknife** deletes each actor in turn; the **parametric** bootstrap
simulates new outcome arrays from the fitted model.

```{r boot}
br <- boot_sir(fit, R = 20, type = "block", seed = 123, trace = FALSE)
br_dyad <- boot_sir(fit, type = "dyad", seed = 123, trace = FALSE)

data.frame(
	method = c("Block Bootstrap", "Dyad Jackknife"),
	valid_refits = c(br$n_valid, br_dyad$n_valid),
	total_refits = c(br$n_total, br_dyad$n_total),
	interval_type = c(br$interval, br_dyad$interval)
)
```

Percentile intervals from block or parametric bootstrap replicates are often more
reliable than Wald intervals for the bilinear parameters:

```{r confint-block}
confint(fit, boot = br)
```

The dyad jackknife uses normal intervals from the jackknife standard error,
because the delete-one fits are not a bootstrap sampling distribution:

```{r confint-dyad}
confint(fit, boot = br_dyad)
```

For reporting, synthesize the uncertainty checks in one table instead of
presenting four disconnected printouts. The rendered version below is a
mechanics check: the block bootstrap used only `R = 20` to keep the vignette
fast. Re-run the same table with `R = 500`--`1000` before treating the block
intervals as evidence.

```{r reporting-table}
ci_classical <- confint(fit, se.type = "classical")
ci_cluster <- confint(fit)
ci_block <- confint(fit, boot = br)
ci_dyad <- confint(fit, boot = br_dyad)

sign_stable <- function(ci) sign(ci[, 1]) == sign(ci[, 2])
interval_pattern <- ifelse(
	sign_stable(ci_cluster) & sign_stable(ci_dyad),
	"same sign in cluster and dyad intervals",
	ifelse(sign_stable(ci_cluster), "same sign in cluster interval only",
	       "interval crosses zero")
)

data.frame(
	term = names(coef(fit)),
	estimate = round(unname(coef(fit)), 3),
	classical = paste0("[", round(ci_classical[, 1], 3), ", ", round(ci_classical[, 2], 3), "]"),
	cluster = paste0("[", round(ci_cluster[, 1], 3), ", ", round(ci_cluster[, 2], 3), "]"),
	block = paste0("[", round(ci_block[, 1], 3), ", ", round(ci_block[, 2], 3), "]"),
	dyad = paste0("[", round(ci_dyad[, 1], 3), ", ", round(ci_dyad[, 2], 3), "]"),
	cluster_over_classical = round(unname(se_cluster / se_classical), 2),
	interval_pattern = interval_pattern,
	block_note = "R = 20; mechanics only",
	row.names = NULL
)
```

Each bootstrap replicate refits the full model; non-converged replicates are
dropped and reported, so check the valid-replicate count. Use `R = 500`--`1000`
for publication-quality block or parametric bootstrap intervals. The dyad
jackknife ignores `R` and instead uses one refit per deleted actor (or sender and
receiver actor for rectangular fits).

## Important caveats on dependence

The classical and HC0-robust SE types assume **independence across dyads**.
Relational data violate this: the $(i, j)$ and $(j, i)$ directed dyads share
members, and degree heterogeneity correlates every tie involving the same actor.
The HC0-robust SEs only correct for heteroskedasticity/overdispersion (note how
little they moved above). The period/block bootstrap resamples whole time
periods as independent blocks: it preserves within-period network dependence,
but not serial dependence across adjacent periods.

The package ships two dyad-aware tools that *do* address it, both shown earlier
in this vignette:

* `vcov()` / `confint()` --- a multiway cluster-robust sandwich (sender +
  receiver + time), the default reporting interval for supported static
  directed-network fits with a stable Hessian.
* `boot_sir(type = "dyad")` --- a delete-one-actor jackknife (also the only
  inference path for full-bilinear bipartite fits).

In practice:

* Report **cluster-robust** SEs (`vcov()`) for supported static fits with
  reliable analytic covariance, and use the bootstrap/jackknife as a
  cross-check.
* For dynamic `W`, unstable Hessians, or full-bilinear bipartite fits, prefer
  the dyad jackknife and response-scale sensitivity checks over cluster SEs.
* Do not use dyad-independent $p$-values as the primary reporting target for
  influence coefficients; report cluster or dyad-jackknife intervals and the
  response-scale scenario table.

## Fixed-receiver model and model comparison

Setting `fix_receiver = TRUE` constrains $B = I$ and estimates only sender-side
influence. This removes the alpha/beta scaling ambiguity by dropping the
receiver-side channel, but the remaining coefficients still need adequate design
rank and signal. It is faster and better-conditioned, and is a useful descriptive
baseline to compare against the full bilinear model on the same response and
mask.

```{r fix-receiver}
fit_fr <- sir(dat$Y, W = dat$W, X = dat$X, Z = dat$Z,
              family = "poisson", fix_receiver = TRUE, calc_se = TRUE, seed = 1)
fr_ci <- confint(fit_fr, se.type = "cluster")
data.frame(
	term = names(coef(fit_fr)),
	estimate = round(unname(coef(fit_fr)), 3),
	cluster_low = round(fr_ci[, 1], 3),
	cluster_high = round(fr_ci[, 2], 3),
	row.names = NULL
)
```

Compare specifications with fit criteria, convergence diagnostics, and a common
response-scale scenario. Here the scenario shifts the first direct covariate by
one standard deviation under each fitted model:

```{r compare}
scenario_change <- function(model) {
	Zcf <- dat$Z
	Zcf[, , 1, ] <- Zcf[, , 1, ] + stats::sd(Zcf[, , 1, ], na.rm = TRUE)
	mu0 <- predict(model)
	mu1 <- predict(model, newdata = list(W = dat$W, X = dat$X, Z = Zcf))
	mean(mu1 - mu0, na.rm = TRUE)
}

data.frame(
	model = c("full bilinear", "fix_receiver"),
	logLik = round(c(as.numeric(logLik(fit)), as.numeric(logLik(fit_fr))), 2),
	AIC   = round(c(AIC(fit), AIC(fit_fr)), 2),
	BIC   = round(c(BIC(fit), BIC(fit_fr)), 2),
	converged = c(fit$convergence, fit_fr$convergence),
	se_reliable = c(fit$se_reliable, fit_fr$se_reliable),
	mean_scenario_change = round(c(scenario_change(fit), scenario_change(fit_fr)), 3)
)
```

Lower AIC/BIC is better, but model choice should also ask whether the simpler
model changes the substantive response-scale conclusion.

## Out-of-sample evaluation

In-sample criteria (AIC/BIC) reward complexity; out-of-sample accuracy is the
more demanding test. `forecast()` produces iterated plug-in forecasts, and
`cv_sir()` runs rolling-origin (expanding-window) cross-validation, scoring each
one-step forecast against the held-out actuals alongside a last-value naive
baseline. `cv_sir()` conditions on the future `Z` and dynamic-`W` slices stored
on the fitted object, so treat its scores as a conditional validation target:
the future covariates must be known, pre-specified, or generated from a scenario
available at the forecast origin.

```{r forecast-cv}
# conditional one-step forecast: fit through T-1, then pass a Z_T slice that is
# known, fixed, or scenario-generated at the forecast origin. In this simulated
# example, the held-out Z_T is used to show the API.
T_hold <- dim(dat$Y)[3] - 1
fit_train <- sir(dat$Y[, , 1:T_hold, drop = FALSE],
                 W = dat$W,
                 X = dat$X[, , 1:T_hold, drop = FALSE],
                 Z = dat$Z[, , , 1:T_hold, drop = FALSE],
                 family = "poisson", fix_receiver = TRUE,
                 calc_se = FALSE, seed = 1)
Z_hold <- dat$Z[, , , T_hold + 1, drop = FALSE]
fc <- forecast(fit_train, h = 1, Z_future = Z_hold)
model_score <- score_sir(dat$Y[, , T_hold + 1, drop = FALSE], fc, "poisson")
naive_score <- score_sir(
	dat$Y[, , T_hold + 1, drop = FALSE],
	array(dat$Y[, , T_hold], dim = dim(fc)),
	"poisson"
)
data.frame(
	metric = names(model_score),
	model = round(unname(model_score), 3),
	naive = round(unname(naive_score), 3),
	improvement = round(unname(naive_score - model_score), 3),
	row.names = NULL
)

# rolling-origin conditional cross-validation: refit on 1..o, score period o + 1
cv <- cv_sir(fit_fr, initial = 60, origins = c(60, 70, 79))
data.frame(
	metric = c("rmse", "mae", "deviance"),
	model = round(unname(cv$aggregate[c("rmse", "mae", "deviance")]), 3),
	naive = round(unname(cv$aggregate[paste0("naive_", c("rmse", "mae", "deviance"))]), 3),
	improvement = round(unname(cv$aggregate[paste0("naive_", c("rmse", "mae", "deviance"))] -
	                           cv$aggregate[c("rmse", "mae", "deviance")]), 3),
	effective_origins = unname(cv$eff_n[c("rmse", "mae", "deviance")]),
	row.names = NULL
)
```

Lower RMSE/MAE/deviance scores are better. This example uses only three origins,
so it demonstrates the workflow rather than providing a final validation study;
with real applications, use more origins or a substantively chosen rolling
schedule.

`score_sir()` exposes the same family-appropriate scores (RMSE / MAE plus the
Poisson deviance) for any observed/predicted pair, so you can assemble custom
backtests.
