---
title: "Getting Started with Social Influence Regression"
author: "Shahryar Minhas & Peter Hoff"
description: "Fit a SIR model, read influence matrices, and use model-implied scenarios on simulated and ICEWS data."
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with Social Influence Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	collapse  = TRUE,
	comment   = "#>",
	fig.align = "center",
	fig.width = 7, fig.height = 5,
	message   = FALSE, warning = TRUE
)
library(sir)
qval <- function(x, prob) unname(stats::quantile(x, prob, na.rm = TRUE))
```

## What problem does SIR solve?

You have a **directed network observed repeatedly over time** --- say, monthly
counts of conflictual events that country $i$ directs at country $j$. Ordinary
regression treats each dyad-time as independent. That misses the defining
feature of relational data: **what an actor does next depends on what the whole
network did last period**, and some actors propagate their behavior onto others
more than others do.

Social Influence Regression (SIR) is designed for that setting. It regresses the
current network $Y_t$ on a lagged, influence-carrying network $X_t$ through a pair of
**$m \times m$ influence matrices**, $A$ and $B$:

$$
  \eta_{i,j,t} = g(\mu_{i,j,t}) \;=\; \boldsymbol{\theta}^{\top} \mathbf{z}_{i,j,t}
  \;+\; \sum_{k,\ell} x_{k,\ell,t}\, a_{i,k}\, b_{j,\ell}
  \;=\; \boldsymbol{\theta}^{\top}\mathbf{z}_{i,j,t} + (A\,X_t\,B^{\top})_{i,j}.
$$

The mean is then $\mu_{i,j,t} = g^{-1}(\eta_{i,j,t})$: expected counts for
Poisson, expected continuous values for Normal, and tie probabilities for
Binomial.

* $A$ is **sender-side** influence: $a_{i,k}$ is how strongly node $k$'s past
  behavior shapes node $i$'s future *outgoing* ties.
* $B$ is **receiver-side** influence: $b_{j,\ell}$ is how strongly node $\ell$
  shapes node $j$'s future *incoming* ties.
* $\boldsymbol{\theta}$ are ordinary **direct effects** of exogenous dyadic
  covariates $\mathbf{z}$.

The key idea (Minhas & Hoff, *Political Analysis* 2025) is that the otherwise
huge $A$ and $B$ are **explained by covariates** $W$:

$$
  A = \sum_{r=1}^{p}\alpha_r W_r, \qquad B = \sum_{r=1}^{p}\beta_r W_r .
$$

So instead of $2m^2$ free influence weights you estimate two short coefficient
vectors $\boldsymbol{\alpha}$ and $\boldsymbol{\beta}$ that say *which observable
features are associated with stronger influence channels*. See
`vignette("methodology")` for the full framework and identification details.

> **Causal interpretation.** In this vignette, "influence" means a lagged,
> model-implied predictive channel. Reading it causally requires a research
> design in which temporal ordering, the exogeneity of `W` and `Z`, the lag
> construction, unmeasured confounding, and the proposed intervention are all
> defensible. Without those assumptions, report SIR estimates as conditional
> temporal associations.

### The three covariate roles (W vs. X vs. Z)

The single most important thing to get straight before fitting is what each input
does:

| Input | Dimension | Role |
|:------|:----------|:-----|
| **Y** | $m \times m \times T$ | outcome network (counts, continuous, or 0/1) |
| **X** | $m \times m \times T$ | the **lagged signal** that flows through influence (typically $\log(Y_{t-1}+1)$ for counts) |
| **W** | $m \times m \times p$ | **influence covariates** that *parameterize* $A$ and $B$ --- what makes a tie a channel of influence (alliance, shared membership, proximity, ...) |
| **Z** | $m \times m \times q \times T$ | **exogenous dyadic covariates** with direct effects $\boldsymbol{\theta}$ (distance, trade, ...) |

In one line: **W describes which dyads can carry lagged influence; X is the
signal that travels; Z are direct effects that bypass the influence machinery.**

## A first fit: simulate, fit, recover

`sim_sir()` generates data from a known SIR process, so you can confirm the
estimator recovers the truth before trusting it on real data. (It bakes in a
$1/(m-1)$ scaling of $X$ so the bilinear predictor stays on a sane scale even as
the network grows; see "Building X" below.)

```{r simulate}
set.seed(42)
dat <- sim_sir(m = 14, T_len = 80, p = 2, q = 2, family = "poisson", seed = 42)

fit <- sir(dat$Y, W = dat$W, X = dat$X, Z = dat$Z, family = "poisson", seed = 1)
fit
```

How well did we recover the truth? `coef()` reports one normalized
parameterization with $\alpha_1$ fixed at 1. That is useful for reading the
software output, but the invariant influence object is
$C = \alpha\beta^\top$ and the fitted means. We start with the normalized
coefficient table because that is what users see from `coef()`:

```{r recovery}
truth     <- c(dat$theta, dat$alpha[-1], dat$beta)
estimated <- coef(fit)

data.frame(
	term      = names(estimated),
	truth     = round(truth, 3),
	estimate  = round(unname(estimated), 3),
	abs_error = round(abs(truth - unname(estimated)), 3)
)
```

The rank-at-most-one influence coefficient matrix is also close:

```{r recovery-c}
C_truth <- outer(dat$alpha, dat$beta)
C_estimated <- outer(fit$alpha, fit$beta)

data.frame(
	quantity = c("maximum absolute C error", "relative Frobenius C error"),
	value = round(c(
		max(abs(C_truth - C_estimated)),
		sqrt(sum((C_truth - C_estimated)^2)) / sqrt(sum(C_truth^2))
	), 3)
)
```

The **point estimates** land close to the truth: direct effects recover tightly
and the influence coefficients are close (noisier per replicate at this network
size, shrinking as you add time periods or nodes). We deliberately show recovery
by *error magnitude* rather than by a Wald "is the truth inside the 95% interval"
check, because the default standard errors are **anti-conservative for network
data** -- they assume independent dyad-times and so understate the true
uncertainty. Treat point recovery as the evidence here, and see
`vignette("sir_inference")` for more defensible uncertainty estimates
(cluster-robust SEs and bootstrap/jackknife checks) before reporting
significance.

> **Aside --- why $\alpha_1 = 1$?** The bilinear term is unchanged if you
> multiply $\boldsymbol{\alpha}$ by a constant $c$ and divide $\boldsymbol{\beta}$
> by $c$. To pin this scale, SIR fixes the first sender coefficient at 1. So
> `coef()` reports $\alpha_2,\dots,\alpha_p$ (each *relative to that baseline of
> 1*) and omits $\alpha_1$, while $\boldsymbol{\beta}$ carries the overall scale.
> That is why the `(alphaW)` block has one fewer row than `(betaW)`.

### Reading the coefficient names

`coef()` and `summary()` tag every parameter by its role:

* `(Z) ...` --- exogenous direct effects ($\boldsymbol{\theta}$)
* `(alphaW) ...` --- sender-influence coefficients ($\alpha_2,\dots,\alpha_p$)
* `(betaW) ...` --- receiver-influence coefficients ($\boldsymbol{\beta}$)

The `...` in each tag is the name of the corresponding `W` slice (so a
covariate named `"ally"` in `dimnames(W)[[3]]` appears as `(alphaW) ally` and
`(betaW) ally`); unnamed slices fall back to `W1`, `W2`, ....

```{r coefficient-table}
ci <- confint(fit)
data.frame(
	term = names(coef(fit)),
	estimate = round(unname(coef(fit)), 3),
	cluster_low = round(ci[, 1], 3),
	cluster_high = round(ci[, 2], 3),
	row.names = NULL
)
```

For applied reporting, use `confint(fit)` or `tidy(fit, conf.int = TRUE)`;
both use cluster-robust uncertainty by default for supported static directed
fits. `summary(fit)` remains a compact Hessian-based printout for quick model
inspection.

## Interpreting influence: who influences whom?

The coefficients say *which covariates* drive influence. To answer *who
influences whom, and by how much*, read the reconstructed influence matrices
`fit$A` and `fit$B`: each off-diagonal `A[i, k]` is node $k$'s influence on node
$i$'s outgoing ties. We blank the diagonal below because self-ties are excluded
from the model's likelihood, so the diagonal of $A$ is not interpreted.

```{r influence-cells}
A <- fit$A
diag(A) <- NA                                      # self-influence is not modeled
ord <- order(abs(A), decreasing = TRUE, na.last = NA)[1:5]
data.frame(
	source_node     = ((ord - 1) %/% nrow(A)) + 1,   # column k (the influencer)
	influenced_node = ((ord - 1) %%  nrow(A)) + 1,   # row i (the influenced)
	influence       = round(A[ord], 3)
)
```

The receiver-side matrix has the same row/column convention, but for incoming
ties: `B[j, l]` says how past activity directed at receiver $l$ shapes future
activity directed at receiver $j$.

```{r receiver-cells}
B <- fit$B
diag(B) <- NA
ord_b <- order(abs(B), decreasing = TRUE, na.last = NA)[1:5]
data.frame(
	source_receiver     = ((ord_b - 1) %/% nrow(B)) + 1,
	influenced_receiver = ((ord_b - 1) %%  nrow(B)) + 1,
	influence           = round(B[ord_b], 3)
)
```

**Reading magnitudes.** The entries of $A$ and $B$ are *unitless relative
influence weights*: their sign and relative size carry the meaning --- a larger
$|a_{i,k}|$ means $k$ shapes $i$ more strongly. The sign of one cell is
conditional on the other bilinear side and on `X`: the contribution is
$A[i,k] X[k,l,t] B[j,l]$. Because influence enters the linear predictor
*bilinearly* through $(A X_t B^{\top})$ with $X$ scaled by $1/(m-1)$, a single
$a_{i,k}$ is **not** a standalone count multiplier. For Poisson direct effects,
$\exp(\theta)$ is a rate ratio and $100(\exp(\theta)-1)$ is the percent change;
for Binomial direct effects it is an odds ratio. For the substantive size of an
influence channel, use a `predict()` scenario (below) rather than exponentiating
a cell. Most off-diagonal weights are small, and many are negative (damping):

```{r influence-magnitude}
round(quantile(A[!is.na(A)], c(0.05, 0.25, 0.5, 0.75, 0.95)), 3)
```

### Visual diagnostics

`plot()` offers six panels: heatmaps of $A$ and $B$ (1--2), their off-diagonal
distributions (3--4), the convergence trace (5), and a coefficient plot (6).
The default `which = 1:4` shows the first four; pass `which = 5:6` for the
convergence and coefficient panels. In the $A$ heatmap **rows are the influenced
node** and **columns are the source**, so a hot cell at (row $i$, column $k$)
means $k$ strongly shapes $i$.

```{r plots, fig.width = 8, fig.height = 8, fig.cap = "SIR diagnostics for the simulated fit: sender and receiver influence heatmaps, followed by off-diagonal influence-weight distributions.", fig.alt = "Four diagnostic panels showing the A and B influence matrix heatmaps and the distributions of their off-diagonal entries."}
plot(fit, which = 1:4)
```

The next chunk draws a network view of the strongest channels when optional
`igraph` and `ggraph` packages are installed; otherwise it prints the same
strongest-channel information as a table.

```{r network, fig.height = 5, fig.cap = "Network view of the strongest fitted sender-side influence channels when igraph and ggraph are installed; otherwise the chunk prints the corresponding top-channel table.", fig.alt = "Directed network plot of sender-side influence channels, with arrows from source nodes to influenced nodes."}
if (requireNamespace("ggraph", quietly = TRUE) &&
	requireNamespace("igraph", quietly = TRUE)) {
	plot_sir_network(fit, matrix = "A", threshold = 0.15)
} else {
	A_network <- fit$A
	diag(A_network) <- NA
	edge_order <- order(abs(A_network), decreasing = TRUE, na.last = NA)[1:5]
	data.frame(
		source_node = ((edge_order - 1) %/% nrow(A_network)) + 1,
		influenced_node = ((edge_order - 1) %% nrow(A_network)) + 1,
		influence = round(A_network[edge_order], 3)
	)
}
```

## Prediction and model-implied scenarios

`predict()` returns fitted values on the response scale. With `newdata` you can
run **model-implied scenarios** --- change one input and hold the rest fixed.
Only `W`, `X`, and `Z` are read; `Y` is never used for prediction. Calling a
scenario a causal counterfactual requires the assumptions in the causal
interpretation box above, plus a coherent way to update the network history and
covariates under the intervention.

```{r predict}
mu_hat <- predict(fit)                             # in-sample expected counts
dim(mu_hat)

Zcf <- dat$Z
Zcf[, , 1, ] <- Zcf[, , 1, ] + sd(Zcf[, , 1, ])    # shift first Z covariate up 1 sd
mu_cf <- predict(fit, newdata = list(W = dat$W, X = dat$X, Z = Zcf))
delta_z <- mu_cf - mu_hat

data.frame(
	scenario = "Increase Z1 by 1 SD",
	baseline_mean = mean(mu_hat, na.rm = TRUE),
	scenario_mean = mean(mu_cf, na.rm = TRUE),
	mean_change = mean(delta_z, na.rm = TRUE),
	median_change = stats::median(delta_z, na.rm = TRUE),
	q05_change = qval(delta_z, 0.05),
	q95_change = qval(delta_z, 0.95),
	check.names = FALSE,
	row.names = NULL
)
```

For influence covariates, the strongest applied workflow is to copy the observed
`W` array and modify a theoretically defined part of it. The helper
`get_scen_array()` builds artificial design grids, which are useful for
diagnostics, but those grids erase the observed network structure. Here we keep
that structure and shift the first continuous `W` slice up or down by one
off-diagonal standard deviation:

```{r scenario-helper}
W_base <- dat$W
off <- row(W_base[, , 1]) != col(W_base[, , 1])
w_step <- stats::sd(W_base[, , 1][off], na.rm = TRUE)

w_scenarios <- lapply(c("-1 SD", "baseline", "+1 SD"), function(label) {
	W_tmp <- W_base
	W1 <- W_tmp[, , 1]
	shift <- switch(label, "-1 SD" = -w_step, "baseline" = 0, "+1 SD" = w_step)
	W1[off] <- W1[off] + shift
	W_tmp[, , 1] <- W1
	mu_tmp <- predict(fit, newdata = list(W = W_tmp, X = dat$X, Z = dat$Z))
	delta <- mu_tmp - mu_hat
	data.frame(
		scenario = label,
		scenario_mean = mean(mu_tmp, na.rm = TRUE),
		mean_change = mean(delta, na.rm = TRUE),
		median_change = stats::median(delta, na.rm = TRUE),
		p90_abs_change = qval(abs(delta), 0.9),
		check.names = FALSE,
		row.names = NULL
	)
})
do.call(rbind, w_scenarios)
```

The table reports expected-count differences on the response scale. A higher
`W` value can move the mean down if the fitted coefficient signs and the
bilinear product imply damping conditional on the observed `X`.

`predict()` returns fitted or scenario-implied expected outcomes. For
one-step-ahead or multi-step forecasting, use `forecast()` so the lagged `X`
updates are handled explicitly.

## A worked example on real data

The package bundles `icews`, a 50-country $\times$ 95-month slice of ICEWS
inter-state material-conflict counts. In this data object, `X` is the raw lagged
count transform `log(Y[,,t-1] + 1)`; it is not divided by $m-1$. To keep the
vignette fast while still showing real output, we fit the first 10 countries and
first 24 months:

```{r icews}
data(icews)
icews_nodes <- 1:10
icews_periods <- 1:24

Y_icews <- icews$Y[icews_nodes, icews_nodes, icews_periods, drop = FALSE]
X_icews <- icews$X[icews_nodes, icews_nodes, icews_periods, drop = FALSE]
W_icews <- icews$W[icews_nodes, icews_nodes, , drop = FALSE]
Z_icews <- icews$Z[icews_nodes, icews_nodes, , icews_periods, drop = FALSE]

data.frame(
	Component = c("Y", "X", "W", "Z"),
	Dimensions = c(
		paste(dim(Y_icews), collapse = " x "),
		paste(dim(X_icews), collapse = " x "),
		paste(dim(W_icews), collapse = " x "),
		paste(dim(Z_icews), collapse = " x ")
	)
)

ifit <- sir(
	Y_icews,
	W = W_icews,
	X = X_icews,
	Z = Z_icews,
	family = "poisson",
	seed = 1,
	max_iter = 20
)
```

Real data is rarely as clean as the simulation. Always check whether the fit is
trustworthy before reading any standard error:

```{r icews-diagnostics}
data.frame(
	Diagnostic = c("ALS converged", "Classical SEs reliable"),
	Value = c(ifit$convergence, ifit$se_reliable)
)
```

For this small subset the Hessian-based SEs are numerically available, but the
cluster/classical gaps below are large enough that classical SEs should not be
read substantively on their own:

```{r icews-coef}
term_labels <- c(
	"(Z) mConf" = "Direct: Lagged Material Conflict",
	"(Z) mConf_ji" = "Direct: Reciprocal Lagged Material Conflict",
	"(Z) minDistLog" = "Direct: Minimum Logged Distance",
	"(Z) ally" = "Direct: Alliance",
	"(Z) verbCoop" = "Direct: Verbal Cooperation",
	"(alphaW) ally" = "Sender Channel: Alliance",
	"(alphaW) verbCoop" = "Sender Channel: Verbal Cooperation",
	"(alphaW) minDistLog" = "Sender Channel: Minimum Logged Distance",
	"(betaW) int" = "Receiver Channel: Baseline",
	"(betaW) ally" = "Receiver Channel: Alliance",
	"(betaW) verbCoop" = "Receiver Channel: Verbal Cooperation",
	"(betaW) minDistLog" = "Receiver Channel: Minimum Logged Distance"
)

se_classical_icews <- sqrt(diag(vcov(ifit, type = "classical")))
se_cluster_icews <- sqrt(diag(vcov(ifit, type = "cluster")))

data.frame(
	"Term" = unname(term_labels[names(coef(ifit))]),
	"Estimate" = round(unname(coef(ifit)), 3),
	"Classical SE" = signif(unname(se_classical_icews), 4),
	"Cluster SE" = signif(unname(se_cluster_icews), 4),
	"Cluster/Classical" = round(unname(se_cluster_icews / se_classical_icews), 2),
	check.names = FALSE
)
```

The direct `Z` rows describe immediate dyadic associations. The `(alphaW)` and
`(betaW)` rows describe which dyadic features are associated with lagged
conflict channels under the fitted SIR model. They are not standalone count-rate
multipliers; use fitted means or scenario predictions for effect sizes. The
first influence covariate, `int`, is an intercept-like baseline channel, so it
plays the role of the fixed $\alpha_1 = 1$.

Country names ride along on `icews$countries`, so you can label the strongest
estimated influence channels. These are exploratory summaries of the fitted
subset, not causal claims:

```{r icews-cells}
A <- ifit$A
dimnames(A) <- list(icews$countries[icews_nodes], icews$countries[icews_nodes])
diag(A) <- NA
top_edges <- order(abs(A), decreasing = TRUE, na.last = NA)[1:5]
source_index <- ((top_edges - 1) %/% nrow(A)) + 1
influenced_index <- ((top_edges - 1) %% nrow(A)) + 1

data.frame(
	Source = colnames(A)[source_index],
	Influenced = rownames(A)[influenced_index],
	Weight = round(A[top_edges], 2)
)
```

The receiver-side channels are read analogously: past activity directed at the
source receiver helps predict future activity directed at the influenced
receiver.

```{r icews-b-cells}
B <- ifit$B
dimnames(B) <- list(icews$countries[icews_nodes], icews$countries[icews_nodes])
diag(B) <- NA
top_b <- order(abs(B), decreasing = TRUE, na.last = NA)[1:5]
b_source_index <- ((top_b - 1) %/% nrow(B)) + 1
b_influenced_index <- ((top_b - 1) %% nrow(B)) + 1

data.frame(
	Source_receiver = colnames(B)[b_source_index],
	Influenced_receiver = rownames(B)[b_influenced_index],
	Weight = round(B[top_b], 2)
)
```

For an effect-size summary, change the lagged signal `X` and compare fitted
means. The next scenario is a post-fit sensitivity check among high-leverage
channels: it nudges the lagged conflict signal sent by the strongest source
country above by one quarter of a standard deviation, holding `W` and `Z` fixed.
It is not a causal intervention or a forecast.

```{r icews-scenario}
mu_icews <- predict(ifit)
X_scen <- X_icews
src <- source_index[1]
x_shift <- 0.25 * stats::sd(X_icews[src, , ], na.rm = TRUE)
X_scen[src, , ] <- X_scen[src, , ] + x_shift
mu_scen <- predict(ifit, newdata = list(W = W_icews, X = X_scen, Z = Z_icews))
delta <- mu_scen - mu_icews

data.frame(
	scenario = paste("Increase lagged signal sent by", colnames(A)[src], "by 0.25 SD"),
	baseline_mean = mean(mu_icews, na.rm = TRUE),
	scenario_mean = mean(mu_scen, na.rm = TRUE),
	mean_change = mean(delta, na.rm = TRUE),
	median_change = stats::median(delta, na.rm = TRUE),
	q05_change = qval(delta, 0.05),
	q95_change = qval(delta, 0.95),
	p90_abs_change = qval(abs(delta), 0.9),
	max_abs_change = max(abs(delta), na.rm = TRUE),
	check.names = FALSE,
	row.names = NULL
)
```

The change columns are expected-count differences on the Poisson response scale.
The median is the main guide to the typical dyad-month; `max_abs_change` tells you
how reactive the fitted model can be in the tail. Because this is a post-fit
sensitivity check on a high-leverage source, inspect the dyad-months that move
most before attaching a substantive interpretation:

```{r icews-scenario-dyads}
top_delta <- order(abs(delta), decreasing = TRUE, na.last = NA)[1:5]
top_idx <- arrayInd(top_delta, dim(delta))
data.frame(
	Source = icews$countries[icews_nodes][top_idx[, 1]],
	Receiver = icews$countries[icews_nodes][top_idx[, 2]],
	Month = icews_periods[top_idx[, 3]],
	baseline_mean = round(mu_icews[top_delta], 3),
	scenario_mean = round(mu_scen[top_delta], 3),
	delta = round(delta[top_delta], 3),
	row.names = NULL
)
```

## Building X (a note that saves silent errors)

`X` is the lagged signal; the right transform depends on the family:

* **Poisson** (counts): `X[,,t] = log(Y[,,t-1] + 1)` --- the log stabilizes the
  multiplicative log link.
* **Normal / Binomial**: the raw lag `X[,,t] = Y[,,t-1]`.

For larger networks, scale `X` by $1/(m-1)$ (as `sim_sir()` does) so the
bilinear term $A X B^{\top}$, which sums over all $m-1$ partners per side, does
not grow with network size and saturate the link. If you forecast from an
unscaled fit such as the bundled `icews` example, pass `infl_scale = 1` or
rebuild `X` with the same scaling you plan to use for future lags.

## Where to next

* `vignette("sir_inference")` --- standard errors, robust SEs, the bootstrap, and
  model comparison.
* `vignette("sir_extensions")` --- Normal/Binomial families, symmetric and
  bipartite networks, dynamic influence covariates, and data-preparation
  utilities.
* `vignette("methodology")` --- the model, identification, estimation algorithms,
  and when to use SIR versus latent-space / AMEN / ERGM.
